---
title: "ETC2420/5242 <br> Multiple linear regression in R"
author: Earo Wang <br> <earo.wang@gmail.com>
date: Lab 06
---

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
opts_chunk$set(warning = FALSE, message = FALSE, 
			   fig.path = 'figure/', cache.path = 'cache/', 
			   fig.align = 'center', dev = 'png', fig.width = 10, fig.height = 6, 
			   fig.show = 'hold', cache = TRUE, external  =  TRUE,
         dev = 'Cairo_png')
read_chunk("lab06.R")
```

# Hands-on data analysis with R

<!-- break -->

```{r library}
```

## Get real data

* Real datasets come in different data files, such as `csv`, `txt`, `json`, `xml`,
  and etc.
    + Getting the current working directory in R

```{r wd}
```

* Read the data in

```{r read, results = "hide"}
```

```{r summary, echo = FALSE}
```

## Poke around the data using graphics

* Look at the correlations using scatter matrix plots

```{r mat, fig.width = 8, fig.height = 8}
```

```{r math}
```

* Look at the distribution of Australian schools

```{r school}
```

## Clean the data

* Making implicit missing data to explicit

```{r na}
```

## Examine each of the variables (pre-modelling)

```{r tally, results = "hide"}
```

I'm lazy... I wrap Di's code into a function and let `purrr::map()` to do the rest.

```{r tally1, results = "hide"}
```

* Actions to take
    + Drop `ST57Q01`
    + Remove missing values

```{r var}
```

Well... a shortcut to remove all the missing cases.

```{r var1}
```

```{r dim}
```

## Transform the data

```{r tran}
```

## Start with linear regression

```{r model}
```

* Heteroscedasticity due to students' socioeconomic and demographic characteristics

```{r hist}
```

## Fit weighted linear regression

```{r model1}
```

```{r inter}
```

# Misc

[Nathaniel's lab help file](http://dicook.github.io/Statistical_Thinking/tutorials/lab06/lab6help.pdf)

## Consultation hours

* Earo
    * **Time**: 13:30 to 15:00 on Thursday
    * **Venue**: Level 11/W1106, Menzies Building
* Nathaniel
    * **Time**: 14:00 to 15:00 on Tuesday
    * **Venue**: Level 11/W1106, Menzies Building

<meta name="copyright" content="LICENSE: CC BY-NC 3.0 US" />
