---
title: "ETC2420/5242 <br> Bootstrapping in R"
author: Earo Wang <br> <earo.wang@gmail.com>
date: Lab 07
---

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
opts_chunk$set(warning = FALSE, message = FALSE, 
			   fig.path = 'figure/', cache.path = 'cache/', 
			   fig.align = 'center', dev = 'png', fig.width = 10, fig.height = 6, 
			   fig.show = 'hold', cache = TRUE, external  =  TRUE,
         dev = 'Cairo_png')
read_chunk("lab07.R")
```

# Happy bootstrapping

<!-- break -->

```{r warmup, echo = FALSE}
```

```{r glm}
```

## Classical confidence intervals

$$
\hat{\beta}_{14} \pm t_{\alpha/2, n - (k + 1)} \times \mathrm{S.E.}(\hat{\beta}_{14})
$$

```{r classical-ci}
```

## Hypothesis test

$$
\begin{align}
H_{0}:& \beta_{14} = 0 \\
H_{1}: & \beta_{14} \neq 0
\end{align}
$$

Compare $t_{crit}$ and $t_{test}$ and decide to reject $H_0$ or not.

## Confidence intervals via bootstrap

1. Make a N boostrap samples (sample data rows, with replacement)
2. Fit the model for each
3. Compute lower and upper C% bounds, by sorting values and pulling the relevant ones, e.g. if N = 1000, C = 95, we would take the 25$^{th}$ and 975$^{th}$ values as the lower and upper CI bounds

```{r q2, eval = FALSE}
```

```{r q2-actual, echo = FALSE}
```

```{r q2-show}
```

## Confidence intervals for predicted value

```{r q3, eval = FALSE}
```

```{r q3-pred, echo = FALSE}
```

Now make a 95% bootstrap confidence interval for predicted value for a new student who is FEMALE, started school at 4, mother and father both work full-time, has a desk, computer and internet, two TVs and 26-100 books in the home. The weight for a student like this is 0.1041. 

```{r q3-data}
```
Be sure to convert the values back into the actual math score range.

## Prediction intervals

1. Compute the residuals from the fitted model 
2. Bootstrap the residuals
3. Find the desired quantiles of the residuals
4. Compute prediction intervals by adding residual quantiles to fitted value

```{r q4-template, eval = FALSE}
```

<meta name="copyright" content="LICENSE: CC BY-NC 3.0 US" />
