---
title: "ETC2420/5242 <br> Testing hypotheses using permutation"
author: Earo Wang <br> & <br> Nathaniel Tomasetti 
date: Lab 03
---

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
opts_chunk$set(warning = FALSE, message = FALSE, 
			   fig.path = 'figure/', cache.path = 'cache/', 
			   fig.align = 'center', dev = 'png', fig.width = 12, fig.height = 7, 
			   fig.show = 'hold', cache = TRUE, external  =  TRUE,
         dev = 'Cairo_png')
read_chunk("lab03.R")
```

# Data structures --- data frame

* Vector
* Matrix
* Array
* List
* **Data frame**
    + `economics`, `gapminder` and `pedestrian` datasets that you looked at in the
      first week are all data frames.
    * More on data structures in R, please read one of the chapters in [Advanced R](http://adv-r.had.co.nz/Data-structures.html).

## Create a data frame

```{r df}
```

```{r glimpse, eval = FALSE}
```

# Make a contingency table

```{r table1, echo = FALSE, results = "hide"}
```

<center>
```{r kable, echo = FALSE}
```
</center>

```{r table1}
```

* However, the R code above is NOT readable, since it needs reading from right to left.
* "Piping" with `%>%` has been introduced to make code more readable.

```{r table2}
```

# Permutation test

```{r prop_dif, eval = FALSE}
```

```{r prop_dif2, echo = FALSE}
```

* Run the function 10000 times
* Set seed to make results reproducible

```{r seed0}
```

```{r diff_prop, eval = FALSE}
```

```{r diff_prop1, echo = FALSE}
```

* Make a histogram

```{r histogram}
```

* Draw a vertical line on the plot that represents the difference for the actual data

```{r vline, echo = FALSE}
```

# Misc

## Online R resources

* [Data wrangling cheet sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
  about `dplyr` and `tidyr`
* [Data visualization cheet sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
  about `ggplot2`
* [R programming course](https://www.coursera.org/learn/r-programming) on Coursera starts this week!

## Consultation hours

* Earo Wang
    * **Time**: 13:30 to 15:00 on Thursday
    * **Venue**: Level 11/W1106, Menzies Building
* Nathaniel Tomasetti
    * **Time**: 14:00 to 15:00 on Tuesday
    * **Venue**: Level 11/W1106, Menzies Building

<meta name="copyright" content="LICENSE: CC BY-NC 3.0 US" />
