---
title: "ETC 2420/5242 Quiz 2"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Name: SOLUTION

Circle your choice.

1. For the normal model density plotted below

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.align='center', fig.width=5, fig.height=3}
library(ggplot2)
library(dplyr)
df <- data.frame(x=seq(-14, 18, 0.2), y=dnorm(seq(-14, 18, 0.2), 2, 4))
ar <- df %>% filter(x<=1) 
ar <- rbind(c(-14, 0), ar, c(1,0))
ggplot(data=df, aes(x=x, y=y)) + geom_line() + 
  geom_polygon(data=ar, aes(x=x, y=y), fill="grey80") + 
  geom_vline(xintercept=1, colour="black") + 
  geom_text(aes(x=1, y=-0.003, label="X"), colour="black", nudge_x = -0.8) +
  geom_text(aes(x=-5, y=0.05, label="Y"), colour="black") +
  geom_segment(aes(x=-1, xend=-4.8, y=0.025, yend=0.048), colour="black") +
  geom_hline(yintercept=0, colour="black") + 
  xlab("") + ylab("") + theme_bw()
```

a. Is the grey area greater than 0.5? `NO`
b. Which letter represents a quantile? `X` 

2. Which of the following would be the density function for a Poisson distribution? (circle it)

$$~~~ P(X=x~|~\lambda) = \frac{\lambda^x e^{-\lambda}}{x!} ~~ x \in \{0, 1, 2, ...\}$$

3. For the following density function, write down the likelihood function for a sample of $n=2, x_1=4, x_2=1$.

$$f(x~|~\lambda) = e^{-\lambda x} ~~ x\geq 0$$

$$ L(\lambda) = e^{-\lambda 4}\times e^{-\lambda }$$

4. For a random variable, $X$, the $P(X>5)=0.3$. What's the probability of observing two sample values (drawn independently from the distribution) greater than 5, in a sample of size 2?

`0.3x0.3=0.09`
