---
title: "ETC 2420/5242 Quiz "
date: "Quiz 5"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  echo=FALSE,
  fig.align='center', 
  fig.height = 4,
  fig.width = 4,
  collapse = TRUE,
  comment = "#>"
)
options(digits=4)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
```

Name: _______________ (Please write your name as it is given in the Monash system, and no need to write your student number.)
Grader: _______________

1. Match the dplyr verb to the action

```{r}
x <- data.frame(verb=c("Arrange", "Summarise", "Filter", "Select", "Mutate", "Ungroup"), action=c("Produce subsets", "Reorder the rows", "Create new variables", "Removing existing grouping", "Choose columns","Make summaries of subsets"))
kable(x)
```

2. In the plot below we are looking at the relationship between humidity and air temperature as measured by Pacific Ocen buoys monitoring El Nino. There are missing values coded as 10% below the minimum on both variables. 

a. Circle the cases that are missing on humidity but not on air temperature.
b. Are the missings on humidity independent of air temperature value?

```{r fig.width=3, fig.height=3}
library(MissingDataGUI)
data(tao)
tao_nomiss <- tao 
tao_nomiss$humidity[is.na(tao_nomiss$humidity)] <-
  min(tao_nomiss$humidity, na.rm=T) - 0.1*diff(range(tao$humidity, na.rm=T))
tao_nomiss$air.temp[is.na(tao_nomiss$air.temp)] <-
  min(tao_nomiss$air.temp, na.rm=T) - 0.1*diff(range(tao$air.temp, na.rm=T))
ggplot(tao_nomiss, aes(x=humidity, y=air.temp)) + geom_point() + theme(aspect.ratio=1)
```

3. Put these words into the right synonym group

`records, attributes, fields, example, instance, features, individuals, subjects, experimental units`

Cases: ______________________________________________________________
Variables: __________________________________________________________

4. For the following data, write down the variables that have been measured.

____________________

```{r}
pew <- read.delim(
  file = "http://stat405.had.co.nz/data/pew.txt",
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = F
)
kable(pew[1:5, 1:5])
```

