---
title: "ETC 2420/5242 Quiz "
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  echo=FALSE,
  fig.align='center', 
  fig.height = 4,
  fig.width = 4,
  collapse = TRUE,
  comment = "#>"
)
options(digits=4)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
```

Name: _______________ (Please write your name as it is given in the Monash system, and no need to write your student number.)
Grader: _______________

1. Which of these two approaches would enforce that the error for each country has the same variance? ($x_1$ has two levels australia, new zealand, $x_2$ is a quantitative variable) A or B ??

A. 

$$y_i = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1*x_2 +\varepsilon$$

B. Fit separate models for austrlia and new zealand 

$$y^1_i = \beta_0 + \beta_1 x_2 + \varepsilon ~~ and ~~y^2_i = \beta_0 + \beta_1 x_2 + \varepsilon $$

2. Prediction intervals for new values are typically wider than confidence intervals for the predicted value? TRUE or FALSE

3. Permutation samples for testing the statistical significance of a slope parameter are consistent with a sample from a population where $\beta_1=0$. TRUE or FALSE

4. Write down the fitted model for this model summary.

```{r}
library(HLMdiag)
radon_lm <- glm(log.radon ~ basement*uranium, data = radon)
summary(radon_lm)
```
