---
title: "ETC 2420/5242 Quiz 1"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Name: SOLUTION

Circle your choice.

1. Is "gender of a baby" typically an example of a random event? True or False

2. Long runs of the same outcome are commonly seen in sequences of random events. True or False

3. For this pair of hypotheses, which is the null hypothesis?

 $$H_o: \mu_1 = \mu_2 ~~~vs~~~ H_a: \mu_1 > \mu_2$$ 
 
4. If the $p$-value is small, e.g. 0.0013, you would typically fail to reject the null hypothesis. True or False
 
5. In the following plot, which of these terms describes the distribution of values:

$$bell-shaped~~~skewed ~left ~~~ skewed ~right ~~~ symmetric$$
$$unimodal~~~ biodal ~~~multimodal ~~~ uniform$$

```{r fig.align='center', fig.width=4, fig.height=2, echo=FALSE, message=FALSE}
library(ggplot2)
df <- data.frame(x=runif(1000), y=rnorm(1000))
ggplot(df, aes(x=x, y=..density..)) + geom_histogram(binwidth=0.1, breaks=seq(0,1,0.1)) +
  scale_x_continuous("", limits=c(0,1)) 
```


