---
title: "ETC 2420/5242 Quiz 3"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Name: _______________ (Please write your name as it is given in the Monash system, and no need to write your student number.)
Grader: _______________

Circle your choice.

1. Sketch the line that corresponds to the following equation. Label and mark your axes appropriately (and no need for a title):

$$y=2x+5$$

\vspace{1cm}

2. For the following fitted model

```{r eval=FALSE, echo=FALSE}
library(readr)
olympics2008 <- read_csv("../data/olympics2008.csv")
olympics2012 <- read_csv("../data/olympics2012.csv")
ol <- merge(olympics2012, olympics2008[,c("Country","Total")],
                     by="Country", all.x=TRUE)
colnames(ol)[7] <- "M2012"
colnames(ol)[12] <- "M2008"

ol_lm <- lm(M2012~M2008, data=ol)
ol_lm
```

$$M_{2012} = 0.74 + 0.96M_{2008}$$

a. What is the value of the intercept for the model?
b. Interpret the slope.
\vspace{1cm}
c. The medal count for Jamaica in 2008 was 12. What would you expect the medal count to be in 2012?
d. The medal count for Indonesia in 2008 was 2 and in 2012 was 5. The predicted medal count for 2012 was 2.66. Compute the residual.
