---
title: 'Statistical Methods for Insurance: Linear Models'
author: Di Cook & Souhaib Ben Taieb, Econometrics and Business Statistics, Monash
  University
date: "W5.C2"
output:
  ioslides_presentation:
    transition: default
    widescreen: yes
  beamer_presentation: default
css: default.css
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  cache = FALSE,
  echo=FALSE,
  fig.height = 2,
  fig.width = 5,
  collapse = TRUE,
  comment = "#>"
)
options(digits=2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
```

## Fit all possible models

Olympic medal tallies:

- Response: M2012
- Explanatory variables: M2008, M2004, Pop, GDP
- Pop and GDP are standardised; Medal counts are on the log10 scale

## Your turn

What are the possible models? How many are there?

## Quick check

```{r fig.align='center', fig.width=8, fig.height=4}
oly_gdp2012 <- read.csv("../data/olympics_gdp_all.csv")
oly_gdp2012_tf <- oly_gdp2012 %>% 
  select(M2012, M2008, M2004, Pop_std, GDP_std)
rownames(oly_gdp2012_tf) <- oly_gdp2012$Country
oly_gdp2012_tf$M2004 <- log10(oly_gdp2012_tf$M2004+1)
oly_gdp2012_tf$M2008 <- log10(oly_gdp2012_tf$M2008+1)
oly_gdp2012_tf$M2012 <- log10(oly_gdp2012_tf$M2012+1)
p1 <- ggplot(oly_gdp2012_tf, aes(x=M2008, y=GDP_std)) + geom_point()
p2 <- ggplot(oly_gdp2012_tf, aes(x=M2008, y=Pop_std)) + geom_point()
library(gridExtra)
grid.arrange(p1, p2, ncol=2)
```

Need to log transform GDP and Pop, too.

##

```{r fig.align='center', fig.width=8, fig.height=4}
oly_gdp2012_tf$GDP_std <- log10(oly_gdp2012_tf$GDP_std+1)
oly_gdp2012_tf$Pop_std <- log10(oly_gdp2012_tf$Pop_std+1)
p1 <- ggplot(oly_gdp2012_tf, aes(x=M2008, y=GDP_std)) + geom_point()
p2 <- ggplot(oly_gdp2012_tf, aes(x=M2008, y=Pop_std)) + geom_point()
grid.arrange(p1, p2, ncol=2)
```

## Fit them - first one

```{r results='hide'}
library(meifly)
oly_gdp2012_tf_models <- fitall(oly_gdp2012_tf[,1], 
                                oly_gdp2012_tf[,-1], "lm")
```

```{r}
summary(oly_gdp2012_tf_models[[1]])
```

## Second one 

```{r}
summary(oly_gdp2012_tf_models[[2]])
```

## Model fit summaries

- Extract the model fit statistics, adjusted $R^2$, $R^2$, AIC, BIC, for each model, and display these against the degrees of freedom
- Flip AIC and BIC, so that the direction matches other statistics
- Helps choose best model
- How different is the best model from the next best model

##

```{r fig.align='center', fig.width=10, fig.height=4}
library(GGally)
library(tidyr)
library(purrr)
library(broom)
oly_gdp2012_tf_models_sum <- oly_gdp2012_tf_models %>%
  map_df(glance) %>%
  mutate(model = 1:15) %>%
  mutate(negBIC = -1*BIC, negAIC = -1*AIC) 
mod_desc <- NULL
for (i in 1:15)
  mod_desc <- c(mod_desc, as.character(summary(oly_gdp2012_tf_models[[i]])$call)[2])
mod_desc <- data.frame(model=1:15, label=mod_desc)
oly_gdp2012_tf_models_sum <- merge(oly_gdp2012_tf_models_sum, 
                                   mod_desc, by="model")

oly_gdp2012_tf_models_sum_m <- oly_gdp2012_tf_models_sum %>%
  gather(fit_stat, val, r.squared, adj.r.squared, logLik, negAIC, 
         negBIC, -BIC, -AIC, -deviance, -df, -label, -model, -statistic, -sigma) %>%
  #group_by(fit_stat) %>% #mutate(stdval = rescale01(val)) %>%
  group_by(fit_stat, df) %>% 
  mutate(rank = min_rank(desc(val)))
oly_gdp2012_tf_models_sum_m$fit_stat <-
  factor(oly_gdp2012_tf_models_sum_m$fit_stat, 
         levels=c("r.squared", "adj.r.squared", "logLik", 
                  "negAIC", "negBIC"))
ggplot(oly_gdp2012_tf_models_sum_m, aes(df, val)) +
  geom_point() +
  geom_line(data=filter(oly_gdp2012_tf_models_sum_m, rank == 1)) +
  facet_wrap(~fit_stat, ncol = 5, scales = "free_y") +
  xlab("Degrees of Freedom") +
  ylab("Values")
```

## What do we learn?

- Not a lot of improvement by adding more variables to the model with one variable 
- Maybe gain a small amount with two variables
- There is a big difference between the best and worst model

## Interactive

```{r fig.align='center', fig.width=5, fig.height=5, fig.show='hold'}
library(plotly)
p1 <- ggplot(oly_gdp2012_tf_models_sum, 
            aes(df, adj.r.squared, label=label)) +
  geom_point() +
  ylim(c(0,1)) +
  xlab("Degrees of Freedom") +
  ylab("Adj R2")
ggplotly(p1)
```

## What do we learn

- The best single variable model uses M2008. The next best single variable model uses M2004. 
- The best two variable model uses M2008 and GDP, but there is very little difference between the next two models (M2008, Pop; M2008, M2004).

##

```{r fig.align='center', fig.width=5, fig.height=5, fig.show='hold'}
library(plotly)
p2 <- ggplot(oly_gdp2012_tf_models_sum, 
            aes(df, logLik, label=label)) +
  geom_point() +
  ylim(c(-50, 35)) + 
  xlab("Degrees of Freedom") +
  ylab("logLik")
ggplotly(p2)
```

##

```{r fig.align='center', fig.width=5, fig.height=5, fig.show='hold'}
p3 <- ggplot(oly_gdp2012_tf_models_sum, 
            aes(df, negBIC, label=label)) +
  geom_point() +
  ylim(c(-110, 50)) +
  xlab("Degrees of Freedom") +
  ylab("-BIC")
ggplotly(p3)
```

## Coefficients for each model

- Extract the parameter estimates for each model 
- Plot these against the explanatory variable name
- Connect values corresponding to the same model with lines

##

```{r fig.align='center', fig.width=8, fig.height=4}
oly_gdp2012_tf_models_coef <- coefficients(oly_gdp2012_tf_models)
oly_gdp2012_tf_models_coef <- merge(oly_gdp2012_tf_models_coef, mod_desc, by="model")
ggplot(data = oly_gdp2012_tf_models_coef, aes(x=variable, y=std)) +
  geom_point() + geom_line(aes(group=model)) +
      ylab("Standardized Estimates") + xlab("Variable")
```

##

```{r fig.align='center', fig.width=10, fig.height=5}
p <- ggplot(data = oly_gdp2012_tf_models_coef, 
            aes(x=variable, y=std, label=label)) +
  geom_point() + geom_line(aes(group=model)) +
      ylab("Standardized Estimates") + xlab("Variable")
ggplotly(p)
```

## What do we learn

- If M2008 is in the model, its coefficient is big and coefficients for other variables are really small. 
- M2004 has a high coefficient, if M2008 is not in the model - it substitutes for M2008.
- GDP and Pop only have high coefficients when M2008 and M2004 are not in the model. 

## Balance complexity vs accuracy

- Choose the simplest model that explains almost as much as a more complex model.

- Choosing here between a single variable model using M2008, and adding GDP. 

## Model A

```{r}
m_A <- lm(M2012~M2008, data=oly_gdp2012_tf)
summary(m_A)
```

## Model B

```{r}
m_B <- lm(M2012~M2008+GDP_std, data=oly_gdp2012_tf)
summary(m_B)
```

## Model comparison 

```{r}
anova(m_A, m_B)
```

## Your turn

How much more explanatory power do we get by including GDP?

## Diagnostics

```{r fig.align='center', fig.width=8, fig.height=4}
m_B_diag <- augment(m_B)
p1 <- ggplot(m_B_diag, aes(x=.fitted, y=M2012)) + geom_point()
p2 <- ggplot(m_B_diag, aes(y=.resid, x=.fitted)) + geom_point()
library(gridExtra)
grid.arrange(p1, p2, ncol=2)
```

## Diagnostics

```{r fig.align='center', fig.width=8, fig.height=4}
m_B_diag <- augment(m_B)
p1 <- ggplot(m_B_diag, aes(x=.hat)) + geom_histogram()
p2 <- ggplot(m_B_diag, aes(x=.cooksd)) + geom_histogram()
grid.arrange(p1, p2, ncol=2)
```

Hmm, do we have a problem with one observation?

##

```{r fig.align='center', fig.width=4, fig.height=4}
p <- ggplot(m_B_diag, aes(x=.fitted, y=.cooksd, label=.rownames)) + geom_point()
ggplotly(p)
```

## VIFs

```{r fig.align='center', fig.width=4, fig.height=4}
library(car)
vif(m_B)
ggplot(m_B_diag, aes(x=M2008, y=GDP_std)) + geom_point()
m_B_coef <- coefficients(m_B)
```

## Final model

- $log10(M2012+1)=$ `r m_B_coef[1]` $+$ `r m_B_coef[2]` $log10(M2008+1)+$ `r m_B_coef[3]` $log10(GDP_{std}+1)+\varepsilon$

- Predict the medal count for Australia $M2008=46$ and $GDP=1507000000000$ ($GDP_{std}=0.3720791$). 

- `r m_B_coef[1]`$+$ `r m_B_coef[2]` $log10(46+1)+$ `r m_B_coef[3]` $log10(0.3720791+1)$ = `r m_B_coef[1]+m_B_coef[2]*log10(46+1)+m_B_coef[3]*log10(0.3720791+1)`

- The observed count for 2012 was 35 medals. What will the estimated medal count be?

## ??

```{r fig.align='center', fig.width=4, fig.height=4}
p1 <- ggplot(m_B_diag, aes(x=.fitted, y=M2012, label=.rownames)) + geom_point()
ggplotly(p1)
```

## Resources

- [Statistics online textbook, Diez, Barr, Cetinkaya-Rundel](https://www.openintro.org/stat/textbook.php?stat_book=isrs)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
