---
title: 'Statistical Methods for Insurance: Randomisation'
author: Di Cook & Souhaib Ben Taieb, Econometrics and Business Statistics, Monash
  University
date: "W1.C1"
output:
  beamer_presentation: default
  ioslides_presentation:
    transition: default
    widescreen: yes
css: default.css
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  cache = FALSE,
  fig.height = 2,
  fig.width = 5,
  collapse = TRUE,
  comment = "#>"
)
options(digits=2)
library(dplyr)
library(tidyr)
```

## Overview of the class

- Topics
- Assessment
- Resources
- Instructors, tutors

## Topics

- Topic 1: Simulation of games for decision strategies (2 weeks))
- Topic 2: Statistical distributions for decision theory (1.5 weeks)
- Topic 3: Linear models for credibility theory (1.5 weeks) 
- Topic 4: Compiling data to problem solve (2 weeks)
- Topic 5: Bayesian statistical thinking (1.5 weeks)
- Topic 6: Temporal data and time series models
- Topic 7: Modeling risk and loss, with data and using randomization to assess uncertainty (2 weeks)

## Assessment

- Final exam: 70%
- Tutorials/labs: 25%,  Weekly reports due Monday noon after the lab
- Class exercises: 5%
- ETC5242 students: Labs 15%, Project report and presentation 10%

## Resources

- Web site: [http://dicook.github.io/Statistical_Thinking/](http://dicook.github.io/Statistical_Thinking/)
- Moodle
- [Statistics online textbook](https://www.openintro.org/stat/textbook.php?stat_book=isrs
)
- [Accuarial online curriculum/exam material](https://www.actuaries.org.uk/studying/plan-my-study-route/fellowshipassociateship/core-technical-subjects/ct6-statistical-methods)
- Software: [R](https://cran.r-project.org), [RStudio Desktop](https://www.rstudio.com/products/rstudio/download2/)

## Instructors

- Instructors: 
    - Professor Di Cook 
    - Dr Souhaib Ben Taieb
- Tutors: 
    - Earo Wang (working with Di on PhD) 
    - Nathaniel Tomasetti (worked with Di for Honors, working with Dr Catherine Forbes on PhD)

## What is randomness?

- Coin flip
- Die roll
- Your sporting team wins
- Gender of a baby
- Rain tomorrow
- Stock price in an hour from now
- Lightning strike
- Pipe burst

## Your turn

We are going to play a game of "Stump the Professor". 

1. Flip a coin. If it shows up tails do A first, if it shows up heads to B first. 

  `A. Write down a sequence of heads and tails that you might expect to come from TEN flips of a coin`
  
  `B. Now flip a coin TEN times, and write down the outcomes`

2. Enter these in the online sheet. Make a note to yourself, whether you did A or B first, but don't tell me. 

3. Now I am going to look at what you entered, and guess if sequence was made up, or actual outcomes from coin flips. 

4. You record how many times I get it right. 

## Example: a look at the Australian electoral distribution

- Results of 2013 election from Australian Electoral Commission web site
- 2011 Census data from the Australian Bureau of Statistics
- Combined demographics of electorate with political representation
- Interactive application, in R package `eechidna`

## How to use randomization to understand probability

```{r echo=FALSE, fig.width=10, fig.height=4}
library(eechidna)
library(dplyr)
library(ggplot2)
aec2013 <- aec2013_2cp_electorate %>%
  filter(Elected == "Y")
aec_abs <- merge(aec2013, abs2011, by = "Electorate")
aec_abs$PartyGp <- aec_abs$PartyAb
aec_abs$PartyGp[aec_abs$PartyGp %in% c("LP","LNP","NP","CLP")] <- "Coalition"
aec_abs$PartyGp[aec_abs$PartyGp %in% c("IND","PUP","KAP","GRN")] <- "Other"
ggplot(data=aec_abs, aes(x=Population)) + geom_dotplot(binwidth=2900) +
  facet_wrap(~PartyGp, ncol = 3) + ylab("") + xlab("Population ('000)") +
  scale_x_continuous(breaks=seq(75000, 225000, 25000), labels=seq(75, 225, 25))
```

## Your turn

- What is the difference (roughly) in population between the biggest and smallest electorates?
- What is the relative worth of a voter in the electorate with the largest population, compared to a voter in the electorate with the smallest population?

## Politics

- Ideally all electorates have exactly the same number of people.
- Geography can interfere with this, e.g an electorate cannot be part in Tasmania and part in Victoria.
- The Australian Electoral Commission will adjust geographic boundaries before each election to adjust for population changes as measured in the most recent Census.

## Compute averages

```{r echo=FALSE}
aec_abs_means <- aec_abs %>% filter(PartyGp != "Other") %>%
  group_by(PartyGp) %>%
  summarise(m = mean(Population, na.rm=T), s = sd(Population, na.rm=T))
aec_abs_means
```

## Statistical thinking

- The means are different
- How big is this difference?
- How likely is this difference to have arisen by chance?

We could use a two-sample t-test to answer these, but here is how to do the equivalent by randomization.

## Procedure

1. Compute the statistic for the data (e.g. absolute value of mean difference)
2. Shuffle the group labels (e.g. put the MP party names into a hat, mix them around, draw them and assign to new electorate)
3. Compute the statistic for this shuffled data
4. Repeat steps 2, 3 many times
5. Examine how often the value of the data statistic, or a larger value occurs

## Let's do it

```{r echo=FALSE, fig.height=4, fig.width=8, fig.align='center'}
library(purrr)
mad <- function(df, shuffle=TRUE) {
  if (shuffle)
    df$PartyGp <- sample(df$PartyGp)
  df_means <- df %>% group_by(PartyGp) %>%
    summarise(m = mean(Population, na.rm=T))
  return(d = abs(df_means$m[1] - df_means$m[2]))
}
aec_abs_sub <- aec_abs %>% filter(PartyGp != "Other")
aec_abs_meandif <- mad(aec_abs_sub, shuffle=FALSE)
aec_abs_shuffle <-1:1000 %>% map_dbl(~ mad(aec_abs_sub))
aec_abs_shuffle <- data.frame(d=aec_abs_shuffle, y=1:1000)
ggplot(data=aec_abs_shuffle, aes(x=d)) + geom_dotplot(binwidth=100) +
  geom_vline(xintercept=aec_abs_meandif, colour="red")
```

Let's also count the number of times that we see a bigger difference by chance. It is `r length(aec_abs_shuffle$d[aec_abs_shuffle$d > aec_abs_meandif])`. 

## What does this mean?

If we oberve a difference this large `r length(aec_abs_shuffle$d[aec_abs_shuffle$d > aec_abs_meandif])` out of `r length(aec_abs_shuffle$d)` random shuffles, is it likely to see this electorate distribution by chance?

## Caveats

Let's wait until the next Census results are in (after August this year) and the latest election results, to compare populations of electorates again. 

## Next steps

- Next few classes we will 
    - review statistical hypothesis testing
    - go through a few more examples of using permutation to make statistical significance tests
    - discuss random number generators, and simulating events
    - compute probabilities of events, and 
- Lab this week will be getting started with using R - the top data science software available today.

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
