---
title: 'Statistical Methods for Insurance: Compiling data for problem solving'
author: Di Cook & Souhaib Ben Taieb, Econometrics and Business Statistics, Monash
  University
date: "W8.C2"
output:
  ioslides_presentation:
    transition: default
    widescreen: yes
  beamer_presentation: default
css: default.css
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  echo=FALSE,
  fig.align='center', 
  fig.height = 4,
  fig.width = 4,
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
```

## Overview of this class

- String operations, working with `text`
- Reading different `data formats`
- Handling missing data

## Working with text

```{r echo=TRUE}
tb <- read_csv("../data/tb.csv")
tb[7:10,1:10]
```

## Convert to long form

```{r echo=TRUE}
tb_long <- tb %>% gather(variable, count, m_04:f_u)
head(tb_long)
```

## String split

```{r echo=TRUE}
tb_long <- tb_long %>% separate(variable, c("gender", "age"), "_")
head(tb_long)
```

## Take a look

```{r echo=TRUE, fig.align='center', fig.width=8, fig.height=4}
tb_long %>% filter(iso2 == "CO", age!="u", year>1990) %>%
  ggplot(aes(x=year, y=count, colour=gender)) + geom_point() + facet_wrap(~age, ncol=3)
```

## Reading different file formats: shapefiles

The Australian Electorate Commission publishes the boundaries of the electorates on their website at [http://www.aec.gov.au/Electorates/gis/gis_datadownload.htm](http://www.aec.gov.au/Electorates/gis/gis_datadownload.htm).

Once the files (preferably the national files) are downloaded, unzip the file (it will build a folder with a set of files). We want to read the shapes contained in the `shp` file into R.

##

```{r echo=TRUE, message=FALSE}
library(maptools)

# shapeFile contains the path to the shp file:
shapeFile <- "../data/vic-esri-24122010/vic 24122010.shp"
sF <- readShapeSpatial(shapeFile)
class(sF)
```

##

`sF` is a spatial data frame containing all of the polygons. 
We use the `rmapshaper` package available from ateucher's github page to thin the polygons while preserving the geography:

```{r echo=TRUE, message=FALSE}
library(rmapshaper)
```

```{r echo=TRUE, message=FALSE}
sFsmall <- ms_simplify(sF, keep=0.05) # use instead of thinnedSpatialPoly
```

##

`keep` indicates the percentage of points we want to keep in the polygons. 5% makes the electorate boundary still quite recognizable, but reduce the overall size of the map considerably, making it faster to plot.

##

We can use base graphics to plot this map:

```{r echo=TRUE, cache=TRUE, message=FALSE, fig.width=6, fig.height=4}
plot(sFsmall)
```

## Extracting the electorate information 

A spatial polygons data frame consists of both a data set with information on each of the entities (in this case, electorates), and a set of polygons for each electorate (sometimes multiple polygons are needed, e.g. if the electorate has islands). We want to extract both of these parts.

##

```{r echo=TRUE, message=FALSE}
nat_data <- sF@data
head(nat_data)
```

##

The row names of the data file are identifiers corresponding to the polygons - we want to make them a separate variable:

```{r echo=TRUE, message=FALSE}
nat_data$id <- row.names(nat_data)
```

## Extracting the polygon information 

The `fortify` function in the `ggplot2` package extracts the polygons into a data frame. 
```{r echo=TRUE}
nat_map <- ggplot2::fortify(sFsmall)
head(nat_map)
```

##

We need to make sure that `group` and `piece` are kept as factor variables - if they are allowed to be converted to numeric values, it messes things up, because as factor levels `9` and `9.0` are distinct, whereas they are not when interpreted as numbers ...

```{r echo=TRUE}
nat_map$group <- paste("g",nat_map$group,sep=".")
nat_map$piece <- paste("p",nat_map$piece,sep=".")
head(nat_map)
```

## Plot it

```{r echo=TRUE, fig.width=4, fig.height=3}
ggplot(nat_map, aes(x=long, y=lat, group=group)) + 
  geom_polygon(fill="white", colour="black") 
```

## Handling missing values

- Need to know how the missings are coded, hopefully clearly missing, treated as NA in R, not 0, or -9, or -9999, or . Recode as need be.

- Study the distribution of missing vs not missing, which will help determine how to handle them.

## What ways can these affect analysis?

- If missings happen when conditions are special, eg sensor tends to stop when temperature drops below 3 degrees Celsius, estimation of model parameters may not reflect the population parameters

- Some techniques, particularly multivariate methods like many used in data mining require complete records over many variables. Just a few missing numbers can mean a lot of cases that cannot be used. 

## Making it Easy - MissingDataGUI

- Methods for summarising missings in a data set

- Ways to plot to examine dependence between missing vs not missing

- Imputation methods to substitute missings

```
library(MissingDataGUI)
data(tao)
MissingDataGUI(tao)
```

## Resources

- [eechidna: Exploring Election and Census Highly Informative Data Nationally
        for Australia](https://cran.r-project.org/web/packages/eechidna/index.html)
- [AEC electorate polygons](http://www.aec.gov.au/Electorates/gis/gis_datadownload.htm)
- [Paper on the MissingDataGUI](https://www.jstatsoft.org/article/view/v068i06/v68i06.pdf)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
