---
title: 'Statistical Methods for Insurance: Generalised Linear Models'
author: Di Cook & Souhaib Ben Taieb, Econometrics and Business Statistics, Monash
  University
date: "W7.C2"
output:
  ioslides_presentation:
    transition: default
    widescreen: yes
  beamer_presentation: default
css: default.css
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  echo=FALSE,
  fig.height = 2,
  fig.width = 5,
  collapse = TRUE,
  comment = "#>"
)
options(digits=2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
library(broom)
```

## Generalised linear models

- Overview
- Types
- Assumptions
- Fitting
- Examples


## Overview

- GLMs are a broad class of models for fitting different types of response variables distributions. 
- The multiple linear regression model is a special case.

## Three components

- Random Component: probability distributin of the response variable
- Systematic Component: explanatory variables
- Link function: describes the relaionship between the random and systematic components

## Multiple linear regression

$$y_i = \beta_0+\beta_1x_1 + \beta_2x_2 + \varepsilon ~~~ or ~~~ E(Y_i)=\beta_0+\beta_1x_1+\beta_2x_2$$

- Random component: $y_i$ has a normal distribution, and so $e_i \sim N(0,\sigma^2)$
- Systematic component: $\beta_0+\beta_1x_1 + \beta_2x_2$
- Link function: identity, just the systematic component

## Poisson regression

$$y_i = exp(\beta_0+\beta_1x_1+\beta_2x_2) + \varepsilon$$

- $y_i$ takes integer values, 0, 1, 2, ...
- Link function: $ln(\mu)$, name=`log`. (Think of $\mu$ as $\hat{y}$.)

## Bernouilli, binomial regression

$$y_i = \frac{exp(\beta_0+\beta_1x_1+\beta_2x_2)}{1+exp(\beta_0+\beta_1x_1+\beta_2x_2)} + \varepsilon$$

- $y_i$ takes integer values, $\{ 0, 1\}$ (bernouilli), $\{ 0, 1, ..., n\}$ (binomial)
- Let $\mu=\frac{exp(\beta_0+\beta_1x_1+\beta_2x_2)}{1+exp(\beta_0+\beta_1x_1+\beta_2x_2)}$, link function is $ln\frac{\mu}{1-\mu}$, name=`logit`

## Assumptions

- The data $y_1, y_2, ..., y_n$ are independently distributed, i.e., cases are independent.
- The dependent variable $y_i$ does NOT need to be normally distributed, but it typically assumes a distribution from an exponential family (e.g. binomial, Poisson, multinomial, normal,...)
- Linear relationship between the transformed response (see examples below)
- Explanatory variables can be transformations of original variables
- Homogeneity of variance does NOT need to be satisfied
- Uses maximum likelihood estimation (MLE) to estimate the parameters
- Goodness-of-fit measures rely on sufficiently large samples

## Example: Olympics medal tally

- Model medal counts on log_GDP
- Medal counts = integer, suggests use a Poisson model. 

```{r fig.align='center', fig.width=6, fig.height=4}
oly_gdp2012 <- read.csv("../data/olympics_gdp_all.csv")
oly_gdp2012 <- oly_gdp2012 %>% 
  mutate(GDP_log=log10(GDP))

ggplot(oly_gdp2012, aes(x=GDP_log, y=M2012)) + geom_point()
```

##

```{r echo=TRUE}
oly_glm <- glm(M2012~GDP_log, data=oly_gdp2012,
               family=poisson(link=log))
```

```{r}
kable(summary(oly_glm)$coefficients)
```

## Your turn

Write down the formula of the fitted model.

## What does this model look like?

```{r fig.align='center', fig.width=8, fig.height=3.5}
oly_augment <- augment(oly_glm) %>%
  mutate(.fitted_exp = exp(.fitted)) %>% 
  arrange(GDP_log)
p1 <- ggplot(oly_gdp2012, aes(x=GDP_log, y=M2012)) + 
  geom_point() +
  geom_line(data=oly_augment, aes(x=GDP_log, y=.fitted_exp),
            colour="red")
p2 <- ggplot(oly_gdp2012, aes(x=GDP_log, y=log(M2012))) + 
  geom_point() +
  geom_line(data=oly_augment, aes(x=GDP_log, y=.fitted),
            colour="red")
grid.arrange(p1, p2, ncol=2)
```

## Example: winning tennis matches

We have data scraped from the web sites of the 2012 Grand Slam tennis tournaments. There are a lot of statistics on matches. Below we have the number of receiving points won, and whether the match was won or not. 

```{r fig.align='center', fig.width=8, fig.height=4}
tennis <- read_csv("../data/tennis_2012_wta.csv")
p1 <- ggplot(tennis, aes(x=Receiving.Points.Won, y=won)) + geom_point()
p2 <- ggplot(tennis, aes(x=Receiving.Points.Won, y=won)) + geom_jitter()
grid.arrange(p1, p2, ncol=2)
```

## Your turn

The response variable is binary. What type of GLM should be fit?

## Model

```{r echo=TRUE}
tennis_glm <- glm(won~Receiving.Points.Won, data=tennis,
                  family=binomial(link='logit'))
```

```{r}
kable(summary(tennis_glm)$coefficients)
```

##

```{r fig.align='center', fig.width=6, fig.height=4}
tennis_augment <- augment(tennis_glm) %>% 
  mutate(.fitted_ln=exp(.fitted)/(1+exp(.fitted))) %>%
  arrange(Receiving.Points.Won)
ggplot(tennis, aes(x=Receiving.Points.Won, y=won)) + geom_point() +
  geom_line(data=tennis_augment, aes(x=Receiving.Points.Won, 
                                y=.fitted_ln), colour="red")
```

## Your turn

Write down the fitted model

## Resources

- [Beginners guide](https://www.analyticsvidhya.com/blog/2015/11/beginners-guide-on-logistic-regression-in-r/)
- [Introduction to GLMs](https://onlinecourses.science.psu.edu/stat504/node/216)
- [Quick-R GLMs](http://www.statmethods.net/advstats/glm.html)
- [The Analysis Factor, Generalized Linear Models Parts 1-4](http://www.theanalysisfactor.com/resources/by-topic/r/)
- [wikipedia](https://en.wikipedia.org/wiki/Generalized_linear_model)
- [Do Smashes Win Matches?](http://onlinelibrary.wiley.com/doi/10.1111/j.1740-9713.2013.00665.x/full)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
