---
title: 'Statistical Methods for Insurance: Multilevel Models'
author: Di Cook & Souhaib Ben Taieb, Econometrics and Business Statistics, Monash
  University
date: "W6.C1"
output:
  ioslides_presentation:
    transition: default
    widescreen: yes
  beamer_presentation: default
css: default.css
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  echo=FALSE,
  fig.height = 4,
  fig.width = 4,
  fig.align='center'
)
options(digits=3)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
library(broom)
```

## Overview of this class

- Fixed effects vs random effects
- Mixed effects models
- Diagnostics

## What is a multilevel model?

- Observations are not independent, but belong to a hierarchy 
- Example: individual level demographics (age, gender), and school level information (location, cours offerings, classroom resources)
- Multilevel model enables fitting different types of dependencies

## Fixed vs random

- `Fixed effects` can be used when you know all the categories, e.g. age, gender, smoking status
- `Random effects` are used when not all groups are captured, and we have a random selection of the groups, e.g. individuals (if you have multiple measurements), schools, hospitals 

## Mixed effects models - a type of multilevel model

For data organized in $g$ groups, consider a continuous response linear mixed-effects model (LME model) for each group $i$, $i=1, \ldots, g$:

$$\underset{(n_i \times 1)}{{\bf y}_i} =  \underset{(n_i \times p)}{{\bf X}_i} \underset{(p \times 1)}{{\boldsymbol \beta}} + \underset{(n_i \times q)}{{\bf Z}_i} \underset{(q \times 1)}{{\bf b}_i} + \underset{(n_i \times 1)}{{\bf \varepsilon}_i}$$

- ${\bf y}_i$ is the vector of outcomes for the $n_i$ level-1 units in group $i$
- ${\bf X}_i$ and ${\bf Z}_i$ are design matrices for the fixed and random effects
- ${\boldsymbol \beta}$ is a vector of $p$ fixed effects governing the global mean structure
- ${\bf b}_i$ is a vector of $q$ random effects for between-group covariance 
- ${\bf \varepsilon}_i$ is a vector of level-1 error terms for within-group covariance

## Example

```{r}
library(HLMdiag)
data(radon)
radon <- radon %>% rename(storey=basement)
```

- Data: $radon$, 919 owner-occupied homes in 85 counties of Minnesota. Available in the `HLMdiag` package
- Response: $log.radon$
- Fixed: $storey$ (categorical), $uranium$ (quantitative)
- Random: $county$ (house is a member of county)

```{r}
glimpse(radon)
```

## Take a look

```{r}
ggplot(radon, aes(x=uranium, y=log.radon)) + geom_point()
```

Plot of response vs covariate. What do you see?

## Here's what we see 

- Vertical stripes: each county is represented by an average uranium value
- Weak linear association, lots of variation for houses within county
- Four points inline horizontally at the base (be suspicious)
- Some counties only have 2, 3 points
- Scales?

## Pre-processing

- Counties with less than 4 observations removed
- Four flat-line observations should be removed, really suspect these were erroneously coded missing values

```{r}
radon_keep <- radon %>% group_by(county) %>% 
  tally() %>% filter(n > 4)
radon_sub <- radon %>% 
  filter(county %in% radon_keep$county & log.radon > -2)
radon_sub$storey <- 
  factor(radon_sub$storey, levels=c(0,1), 
         labels=c("basement", "first floor"))
```

## Look again

```{r fig.width=8}
ggplot(radon_sub, aes(x=uranium, y=log.radon)) + 
  geom_point() +
  geom_smooth(method="lm", se=F) +
  facet_wrap(~storey)
```

## Fit a simple model

$$log.radon = \beta_0 + \beta_1 storey + \beta_2 uranium + \varepsilon$$

```{r}
radon_lm <- glm(log.radon ~ storey + uranium, 
                data = radon_sub)
summary(radon_lm)
```

## Your turn

Make a sketch of what this model looks like.

```{r echo=FALSE, eval=FALSE}
radon_lm_fit <- radon_sub; radon_lm_fit$fit <- fitted(radon_lm)
ggplot(radon_lm_fit, aes(x=uranium, y=log.radon, colour=storey)) + geom_point() + 
  geom_line(aes(y=fit))
```

## Fit an interaction term

```{r}
radon_lm <- glm(log.radon ~ storey*uranium, data = radon_sub)
summary(radon_lm)
radon_lm_fit <- radon_sub; radon_lm_fit$fit <- fitted(radon_lm)
```

## What does this model look like?

```{r fig.width=6}
ggplot(radon_lm_fit, aes(x=uranium, y=log.radon, colour=storey)) +
  geom_point(alpha=0.3) + 
  geom_line(aes(y=fit), size=1)
```

## Your turn

Write down the equation of the fitted model

## Mixed effects model

$$log.radon_{ij} = \beta_0 + \beta_1 storey_{ij} + \beta_2 uranium_i + b_{0i} + b_{1i} storey_{ij} + \varepsilon_{ij}$$

$$~~~ i=1, ..., \#counties; j=1, ..., n_i$$

```{r echo=TRUE, results='hide'}
library(lme4)
radon_lmer <- lmer(log.radon ~ storey + uranium + 
  (storey | county.name), data = radon_sub)
summary(radon_lmer)
radon_lmer_fit <- augment(radon_lmer) 
```

## Your turn

For the radon data:

- What is $p$, $q$, $g$?
- And hence $n_i, i=1, \dots, g$?

$$log.radon_{ij} = \beta_0 + \beta_1 storey_{ij} + \beta_2 uranium_i + b_{0i} + b_{1i} storey_{ij} + \varepsilon_{ij}$$

$$~~~ i=1, ..., \#counties; j=1, ..., n_i$$

```{r eval=FALSE, echo=FALSE}
# p=2 (storey, uranium)
# q=1 (storey)
length(unique(radon_sub$county.name)) # g
radon_sub %>% group_by(county.name) %>% tally()
```

## Examining the model output: fixed effects

```
Fixed effects:
                    Estimate Std. Error t value
(Intercept)          1.48066    0.03856   38.40
storeyfirst floor   -0.59011    0.11246   -5.25
uranium              0.84600    0.09532    8.88
```

How do these compare with the simple linear model estimates?

```
Coefficients:
                  Estimate Std. Error t value Pr(>|t|)    
(Intercept)        1.44830    0.03131  46.254  < 2e-16 ***
storeyfirst floor -0.61125    0.07332  -8.337 3.35e-16 ***
uranium            0.83591    0.07422  11.262  < 2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

## Examining the model output: random effects

```
Random effects:
 Groups      Name                Variance Std.Dev. Corr
 county.name (Intercept)         0.01388  0.1178       
             storeyfirst floor   0.22941  0.4790   0.02
 Residual                        0.50694  0.7120       
Number of obs: 796, groups:  county.name, 46
```

This is saying that the variance of the estimates for first floor observations is larger than the storey. 

## What it looks like

```{r fig.width=7}
ggplot(radon_lmer_fit, aes(x=uranium, y=log.radon)) + 
  geom_point(alpha=0.2) + 
  geom_point(aes(y=.fitted, colour=county.name)) + 
  facet_wrap(~storey) + theme(legend.position="none")
```

## Or like this

```{r fig.width=6}
ggplot(radon_lmer_fit, aes(x=uranium, y=log.radon)) + 
  geom_point(alpha=0.2) + 
  geom_point(aes(y=.fitted, colour=storey)) 
```

## Your turn

How does the mixed effects model differ from the simple linear model? (Hint: Think about the variance.)

```{r fig.width=6}
ggplot(radon_lm_fit, aes(x=uranium, y=log.radon, colour=storey)) +
  geom_point(alpha=0.3) + 
  geom_line(aes(y=fit), size=1)
```

## Assumptions

Recall:

 $$\underset{(n_i \times 1)}{{\bf y}_i} =  \underset{(n_i \times p)}{{\bf X}_i} \underset{(p \times 1)}{{\boldsymbol \beta}} + \underset{(n_i \times q)}{{\bf Z}_i} \underset{(q \times 1)}{{\bf b}_i} + \underset{(n_i \times 1)}{{\boldsymbol \varepsilon}_i}$$

- ${\bf b}_i$ is a random sample from $\mathcal{N}({\bf 0}, {\bf D})$ and independent from the level-1 error terms,  
- ${\boldsymbol \varepsilon}_i$ follow a $\mathcal{N}({\bf 0},\sigma^2 {\bf R}_i)$ distribution 
- ${\bf D}$ is a positive-definite $q \times q$ covariance matrix and ${\bf R}_i$ is a positive-definite $n_i \times n_i$ covariance matrix

## Extract and examine level-1 residuals

```{r fig.show='hide'}
radon_lmer_fit$resid1 <-  HLMresid(radon_lmer, 
           level=1)
ggplot(radon_lmer_fit, aes(x=resid1)) + 
  geom_histogram(binwidth=0.5) 
```

${\boldsymbol \varepsilon}_i \sim \mathcal{N}({\bf 0},\sigma^2 {\bf R}_i)$

```{r echo=FALSE}
ggplot(radon_lmer_fit, aes(x=resid1)) + 
  geom_histogram(binwidth=0.5) 
```

Level-1 (observation level) look normal. 

## qqplot

```{r}
ggplot_qqnorm(radon_lmer_fit$resid1, line="rlm") +
  theme(aspect.ratio=1)
```

Level-1 (observation level) do look nearly normal. 

## Examine within group

Summary statistics

```{r}
radon_lmer_fit %>% group_by(county.name) %>%
  summarise(m = mean(resid1), s = sd(resid1), n = length(resid1)) %>%
  head()
```

##

```{r echo=FALSE, fig.height=6}
res.sum <- radon_lmer_fit %>% group_by(county.name) %>%
  summarise(m = mean(resid1), s = sd(resid1), n = length(resid1))
ord <- order(res.sum$m)
radon_lmer_fit$county.name <- factor(radon_lmer_fit$county.name, levels=res.sum$county.name[ord])
ggplot(radon_lmer_fit, aes(x=county.name, y=resid1)) + 
  geom_point(alpha=0.5) + coord_flip()
```

## Learn

There is some difference on average between counties, which means that residuals still have some structure related to the county location.

## Normality tests

Anderson-Darling, Cramer-von Mises, Lilliefors (Kolmogorov-Smirnov)

```{r results='hide'}
library("nortest")
ad.test(radon_lmer_fit$resid1)
cvm.test(radon_lmer_fit$resid1)
lillie.test(radon_lmer_fit$resid1)
```

```{r echo=FALSE}
ad.test(radon_lmer_fit$resid1)
```

all believe that the residuals are consistent with normality.

## Conclusion about level-1 residuals

The assumption: 

$${\boldsymbol \varepsilon}_i \sim \mathcal{N}({\bf 0},\sigma^2 {\bf R}_i)$$

is probably ok, at the worst it is not badly violated. 

## Random effects

$${\bf b}_i \sim \mathcal{N}({\bf 0}, {\bf D}), ~~~ i=1, \dots g$$

where ${\bf D}$ allows for correlation between random effects within group, and these should be independent from the level-1 error

```{r}
rf <- HLMresid(radon_lmer, level="county.name") 
# same as ranef(radon_lmer)
rf$county.name <- rownames(rf)
rf <- rf %>% rename(resid.basement=`(Intercept)`, 
                    resid.ff=`storeyfirst floor`)
radon_lmer_fit <- merge(radon_lmer_fit, rf, 
                        by="county.name")
```

We have both intercepts (basement) and slopes (first floor)

##

```{r fig.width=8}
p1 <- ggplot(radon_lmer_fit, aes(x=resid.basement)) + 
  geom_histogram(binwidth=0.05) + ggtitle("basement")
p2 <- ggplot(radon_lmer_fit, aes(x=resid.ff)) + 
  geom_histogram(binwidth=0.2) + ggtitle("first floor")
grid.arrange(p1, p2, ncol=2)
```

##

```{r fig.width=8}
p1 <- ggplot_qqnorm(radon_lmer_fit$resid.basement, line="rlm") +
  theme(aspect.ratio=1) + ggtitle("basement")
p2 <- ggplot_qqnorm(radon_lmer_fit$resid.ff, line="rlm") +
  theme(aspect.ratio=1) + ggtitle("first floor")
grid.arrange(p1, p2, ncol=2)
```

## Should be no correlation 

```{r}
ggplot(radon_lmer_fit, aes(x=resid.basement, y=resid.ff)) + 
  geom_point() + theme(aspect.ratio=1) 
```

## Fitted vs Observed

Plotting the observed vs fitted values, gives a sense for how much of the response is explained by the model. Here we can see that there is still a lot of unexplained variation.

```{r echo=FALSE}
ggplot(radon_lmer_fit, aes(x=.fitted, y=log.radon)) + 
  geom_point()  
```

## Goodness of fit

From the linear model

```{r}
glance(radon_lm)
```

From the random effects model

```{r}
glance(radon_lmer)
```

Hmmm... deviance looks strange! Compute sum of squares of residuals instead:

```{r}
sum(radon_lmer_fit$resid1^2)
```

Which model is best?

## Influence

```{r}
ggplot(radon_lmer_fit, aes(x=.fitted, y=.cooksd)) + geom_point()
```

No overly influential observations

## Resources

- [HLMDiag package explanation](https://www.jstatsoft.org/article/view/v056i05)
- [HLM package](https://cran.r-project.org/web/packages/HLMdiag/index.html)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
