---
title: 'Statistical Methods for Insurance: Statistical distributions'
author: Di Cook & Souhaib Ben Taieb, Econometrics and Business Statistics, Monash
  University
date: "W3.C2"
output:
  ioslides_presentation:
    transition: default
    widescreen: yes
  beamer_presentation: default
css: default.css
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning=FALSE,
  error=FALSE,
  cache = FALSE,
  echo=FALSE,
  fig.height = 2,
  fig.width = 5,
  collapse = TRUE,
  comment = "#>"
)
options(digits=2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
```

## Overview of this class

- Random variables
- Central limit theorem
- Estimation
- Quantiles
- Goodness of fit
- READING: CT6, Section 1.3-1.9

## Random variables vs random samples

- Conceptually we think about a **random variable** $(X)$ having a distribution

$$X \sim N(\mu, \sigma)$$

- Once we collect data, or use simulation we will have a realisation from the distribution, a random sample, observed values:

```{r}
x <- rnorm(5, 0, 1)
x
```

```{r fig.align='center', fig.width=3, fig.height=2, fig.show='hold'}
df <- data.frame(x=seq(-4,4, 0.008),  d=dnorm(seq(-4,4, 0.008)))
df2 <- data.frame(x=x, y=0)
ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + 
  geom_point(data=df2, aes(x=x, y=0), colour="red")
```

## Central limit theorem

- Why the normal model is so fundamental 
- Regardless what distribution $X$ has, the mean of a sample will have a normal distribution, if the sample size is large:

"Let $\{X_1, ..., X_n\}$ be a random sample of size $n$ — that is, a sequence of independent and identically distributed random variables drawn from a distribution mean given by $\mu$ and finite variance given by $\sigma^2$. The sample average is defined $\bar{X} = \sum_{i=1}^{n} X_i/n$, then as $n$ gets large the distribution of $\bar{X}$ approximates $N(\mu, \sigma/\sqrt{n})$."

## Example: Uniform parent

```{r fig.align='center', fig.width=8, fig.height=5}
rs <- runif(10000000)
x2 <- apply(matrix(rs, ncol=2), 1, mean)
x20 <- apply(matrix(rs, ncol=20), 1, mean)
x50 <- apply(matrix(rs, ncol=50), 1, mean)
x100 <- apply(matrix(rs, ncol=100), 1, mean)
df <- data.frame(pop=rs[1:1000], x2=x2[1:1000], x20=x20[1:1000], 
                 x50=x50[1:1000], x100=x100[1:1000]) 
p1 <- ggplot(data.frame(rs), aes(x=rs)) + geom_histogram(breaks=seq(0,1,0.1)) +
  ggtitle("POPULATION") + xlab("")
p2 <- ggplot(df, aes(x=x2)) + geom_histogram(binwidth=0.1) +
  ggtitle("n=2") + xlab("") + xlim(c(0,1))
p3 <- ggplot(df, aes(x=x20)) + geom_histogram(binwidth=0.05) +
  ggtitle("n=20") + xlab("") + xlim(c(0,1))
p4 <- ggplot(df, aes(x=x100)) + geom_histogram(binwidth=0.02) +
  ggtitle("n=100") + xlab("") + xlim(c(0,1))
grid.arrange(p1, p2, p3, p4, ncol=2)
```

## Example: Weibull parent

```{r fig.align='center', fig.width=8, fig.height=5}
rs <- rweibull(1000000, 1.5)
x2 <- apply(matrix(rs, ncol=2), 1, mean)
x20 <- apply(matrix(rs, ncol=20), 1, mean)
x50 <- apply(matrix(rs, ncol=50), 1, mean)
x100 <- apply(matrix(rs, ncol=100), 1, mean)
df <- data.frame(pop=rs[1:1000], x2=x2[1:1000], x20=x20[1:1000], 
                 x50=x50[1:1000], x100=x100[1:1000]) 
p1 <- ggplot(data.frame(rs), aes(x=rs)) + geom_histogram() +
  ggtitle("POPULATION") + xlab("") + xlim(c(0,6))
p2 <- ggplot(df, aes(x=x2)) + geom_histogram(binwidth=0.2) +
  ggtitle("n=2") + xlab("") + xlim(c(0,6))
p3 <- ggplot(df, aes(x=x20)) + geom_histogram(binwidth=0.1) +
  ggtitle("n=20") + xlab("")  + xlim(c(0,6))
p4 <- ggplot(df, aes(x=x100)) + geom_histogram(binwidth=0.05) +
  ggtitle("n=100") + xlab("")  + xlim(c(0,6))
grid.arrange(p1, p2, p3, p4, ncol=2)
```

## Estimation

- Estimate parameters of a distribution from the sample data
- Common approach is maximum likelihood estimation
- Requires assuming we know the basic functional form 

## Maximum likelihood estimate (MLE)

- Estimate the unknown parameter $\theta$ using the value that maximises the probability (i.e. likelihood) of getting the observed sample
- Likelihood function

$$
\begin{aligned}
L(\theta) &=P(X_1=x_1,X_2=x_2, ... ,X_n=x_n~|~\theta) \\      &=f(x_1|\theta)f(x_2|\theta)\cdots f(x_n|\theta) \\
&=\prod_{i=1}^n f(x_i;\theta)
\end{aligned}
$$

- This is now a function of $\theta$.
- Use function maximisation to solve.

## Example - Mean of normal distribution, assume variance is 1

- MLE estimate of the population mean for a normal model is the sample mean
- Run this numerically
- Suppose we have a sample of two: $x_1=1.0, x_2=3.0$
- Likelihood

$$
\begin{aligned}
L(\mu) &= \frac{1}{\sqrt{2\pi}}e^{-\frac{(1.0-\mu)^2}{2}}\times
            \frac{1}{\sqrt{2\pi}}e^{-\frac{(3.0-\mu)^2}{2}}\\
      &= \frac{1}{2\pi}e^{-\frac{(1-\mu)^2+(3-\mu)^2}{2}}
\end{aligned}
$$     

## Plot it

```{r fig.align='center', fig.width=5, fig.height=3}
nmle <- function(mu) {
  f <- 1/(2*pi) * exp(-((1-mu)^2+(3-mu)^2)/2)
  return(f)
}
df <- data.frame(x=seq(1, 3, 0.1), f=nmle(seq(1, 3, 0.1)))
ggplot(df, aes(x=x, y=f)) + geom_line() + xlab(expression(mu)) + ylab("L")
```

- The maximum is at 2.0. This is the sample mean, which we can prove algebraically is the MLE. 

## Estimate mean and variance

Sample

```{r}
set.seed(1031)
x <- rnorm(22, 3, 5)
x
```

We know it comes from a normal distribution. What are the best guesses for the $\mu, \sigma$?

##

Compute the likelihood for a range of values of both parameters. 

```{r fig.align='center', fig.width=5, fig.height=5}
nmle <- function(x, mu, sigma) {
  f <- prod(dnorm(x, mu, sigma))
  return(f)
}
mu <- seq(1, 5, 0.1)
sig <- seq(2, 7, 0.1)
g <- expand.grid(x=mu, y=sig)
g$f <- 0
for (i in 1:nrow(g)) {
  g$f[i] <- nmle(x, g$x[i], g$y[i])
}
ggplot(g, aes(x=x, y=y, fill=f)) + geom_tile() + xlab(expression(mu)) + ylab(expression(sigma)) + theme_bw() + 
  scale_fill_continuous("L") +
  theme(aspect.ratio=1)
```


## Quantiles

- `quantiles` are cutpoints dividing the range of a probability distribution into contiguous intervals with equal probabilities
- 2-quantile is the median (divides the population into two equal halves)
- 4-quantile are quartiles, $Q_1, Q_2, Q_3$, dividing the population into four equal chunks
- quantiles are values of the random variable $X$
- useful for comparing distributions

## Example: 

- 12-quantiles for a $N(0,1)$

```{r echo=TRUE}
qnorm(seq(1/12,11/12,1/12))
```

- 23-quantiles from a $Gamma(2,1)$

```{r echo=TRUE}
qgamma(seq(1/23,22/23,1/23), 2)
```

## Percentiles

- indicate the value of $X$ below which a given percentage of observations fall, e.g. 20th percentile is the value that has 20% of values below it
- 17th percentile from $N(0,1)$

```{r echo=TRUE}
qnorm(0.17)
```

- 78th percentile from $Gamma(2,1)$

```{r echo=TRUE}
qgamma(0.78, 2)
```

## Goodness of fit

- Quantile-quantile plot (QQplot) plots theoretical vs sample quantiles
- Lets check the distribution of PISA math scores

```{r fig.align='center', fig.width=8, fig.height=4}
st <- readRDS("../data/student_sub.rds")
aus <- st %>% filter(CNT == "AUS")
n <- nrow(aus)
aus_math <- data.frame(PV1MATH = sort(scale(aus$PV1MATH)), 
  q=qnorm(c(1 - 0.5^(1/n), (2:(n-1) - 0.3175) / (n + 0.365),0.5^(1/n))))
p1 <- ggplot(aus_math, aes(x=PV1MATH)) + geom_histogram(binwidth=0.5) +
  xlab("Standardised math score")
p2 <- ggplot(aus_math, aes(x=PV1MATH, y=q)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  xlab("Sample quantiles") + ylab("Theoretical quantile") +
  theme(aspect.ratio=1)
grid.arrange(p1, p2, ncol=2)
```

## QQ-Plot computation

1. Sort and standardise the sample values from low to high
2. Theoretical quantiles, $n=$ sample size
$$  1 - 0.5^{(1/n)} ~~~ i=1 \\
   ~~~~~~~~~~~~~\frac{i - 0.3175}{n + 0.365} ~~~ i=2, ..., n-1 \\
   0.5^{(1/n)} ~~~~~~ ~~~  i=n $$
3. Plot the theoretical vs sample quantiles 

## Reading QQ-plots

- The points should lie along the $X=Y$ line, for the sample to be consistent with the distribution.
- How close is good enough?
- It depends on the sample size.
- Simulate some samples of the same size from the target distribution, and make QQ-plots of these, to compare with the actual data

## 

```{r fig.align='center', fig.width=8, fig.height=5}
aus_math <- data.frame(PV1MATH = sort(scale(aus$PV1MATH)), 
  q=qnorm(c(1 - 0.5^(1/n), (2:(n-1) - 0.3175) / (n + 0.365),0.5^(1/n))), 
  s1=sort(rnorm(n)), s2=sort(rnorm(n)), s3=sort(rnorm(n)), 
  s4=sort(rnorm(n)), s5=sort(rnorm(n)))
p1 <- ggplot(aus_math, aes(x=PV1MATH, y=q)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  xlab("Sample quantiles") + ylab("Theoretical quantile") +
  theme(aspect.ratio=1) + ggtitle("(a)")
p2 <- ggplot(aus_math, aes(x=s1, y=q)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  xlab("Sample quantiles") + ylab("Theoretical quantile") +
  theme(aspect.ratio=1) + ggtitle("(b)")
p3 <- ggplot(aus_math, aes(x=s2, y=q)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  xlab("Sample quantiles") + ylab("Theoretical quantile") +
  theme(aspect.ratio=1) + ggtitle("(c)")
p4 <- ggplot(aus_math, aes(x=s3, y=q)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  xlab("Sample quantiles") + ylab("Theoretical quantile") +
  theme(aspect.ratio=1) + ggtitle("(d)")
p5 <- ggplot(aus_math, aes(x=s4, y=q)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  xlab("Sample quantiles") + ylab("Theoretical quantile") +
  theme(aspect.ratio=1) + ggtitle("(e)")
p6 <- ggplot(aus_math, aes(x=s5, y=q)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  xlab("Sample quantiles") + ylab("Theoretical quantile") +
  theme(aspect.ratio=1) + ggtitle("(f)")
grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3)
```

## How different is exponential from Pareto?

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
rpareto <- function(n, c){
 if (c<=0) stop("c must be positive")
   rp <- runif(n)^(-1/c)
  rp
}
df <- data.frame(re=sort(scale(rexp(1000))), rp=sort(scale(rpareto(1000, 1))))
p1 <- ggplot(df, aes(x=re)) + geom_histogram() + ggtitle("Exp")
p2 <- ggplot(df, aes(x=rp)) + geom_histogram() + ggtitle("Pareto")
p3 <- ggplot(df, aes(x=re, y=rp)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() +
  xlab("exp") + ylab("Pareto") + ggtitle("QQ-plot") + theme(aspect.ratio=1)
grid.arrange(p1, p2, p3, ncol=3)
```

## How different can exponentials be?

```{r fig.align='center', fig.width=8, fig.height=5, fig.show='hold'}
df <- data.frame(re=sort(scale(rexp(100))), rp=sort(scale(rpareto(100, 1))),
                 re2=sort(scale(rexp(100))), re3=sort(scale(rexp(100))),
                 re4=sort(scale(rexp(100))), re5=sort(scale(rexp(100))))
p1 <- ggplot(df, aes(x=re, y=rp)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlab("") + ylab("") +
  theme(aspect.ratio=1) + ggtitle("(a)")
p2 <- ggplot(df, aes(x=re, y=re2)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlab("") + ylab("") +
  theme(aspect.ratio=1) + ggtitle("(b)")
p3 <- ggplot(df, aes(x=re, y=re3)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlab("") + ylab("") +
  theme(aspect.ratio=1) + ggtitle("(c)")
p4 <- ggplot(df, aes(x=re, y=re3)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlab("") + ylab("") +
  theme(aspect.ratio=1) + ggtitle("(d)")
p5 <- ggplot(df, aes(x=re, y=re4)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlab("") + ylab("") +
  theme(aspect.ratio=1) + ggtitle("(e)")
p6 <- ggplot(df, aes(x=re, y=re5)) + 
  geom_abline(intercept=0, slope=1) +
  geom_point() + xlab("") + ylab("") +
  theme(aspect.ratio=1) + ggtitle("(f)")
grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3)
```

## Resources

- [wikipedia](https://en.wikipedia.org/wiki/Quantile)
- [PSU 414](https://onlinecourses.science.psu.edu/stat414/node/191)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
