---
title: 'Statistical Methods for Insurance: Statistical distributions'
author: Di Cook & Souhaib Ben Taieb, Econometrics and Business Statistics, Monash
  University
date: "W3.C1"
output:
  ioslides_presentation:
    transition: default
    widescreen: yes
  beamer_presentation: default
css: default.css
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  echo=FALSE,
  fig.height = 2,
  fig.width = 5,
  collapse = TRUE,
  comment = "#>"
)
options(digits=2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(gridExtra)
```

## Overview of this class

- Quiz 1 solution
- Random numbers
- Mapping random numbers to events for simulation
- Statistical distributions
- READING: CT6, Section 1.3-1.9

## Random numbers

- True random number generators: [Radioactive decay](https://www.fourmilab.ch/hotbits/), [electromagnetic field of a vacuum](https://qrng.anu.edu.au)
- Computers only technically provide pseudo-random numbers, using deterministic process, e.g linear congruential, for large $a, b, m$

$$X_{n+1} = (aX_n + b) ~~mod ~~m$$

## RANDU - a bad PRNG

- Used in the 60s and onwards

$$X_{n+1} = 65539 X_n ~~mod ~~2^{31}$$

```{r fig.width=3.5, fig.height=3.5, fig.align='center'}
randu <- read.csv("randu.csv")
ggplot(randu, aes(x=V1, y=V2)) + geom_point() + theme(aspect.ratio=1)
```

## Mersenne Twister

- algorithm is a twisted generalised feedback shift register (TGFSR)
- based on a Marsenne prime, $2^m-1$
- most commonly used today
- each integer will occur the same number of times in a period

## Using random numbers

- Random number tables deliver single digits 0, 1, ..., 9
- When using these you need to ensure that you map these digits or combinations of the digits to match the probabilities of events
- For example, use random numbers to sample students from class
    + There are 105 students in the class
    + Need to use three sequential digits
    + BUT there are 1000 three digit numbers, so either we will throw away 895 of them, or we could map a person to multiple numbers (9) and throw away only 55
    + If any person is selected more than once, throw out repeats

## Estimate the proportion of 2420:5242

Class list:

```{r}
classlist <- read_csv("../labs/groups-ETC2420 - ETC5242 - S2 2016-20160809120717.csv")
sample_class <- classlist %>% 
  separate(`Full name`, c("First", "Last"), sep=" ") %>%
  arrange(Last) %>% select(First, Last) 
sample_class_sub <- sample_class[1:5,]
sample_class_sub$numbers <- c("001,002,003,004,005,006,007,008,009",
                              "010,011,012,013,014,015,016,017,018,019",
                              "020,021,022,023,024,025,026,027,028,029",
                              "030,031,032,033,034,035,036,037,038,039",
                              "040,041,042,043,044,045,046,047,048,049")
kable(sample_class_sub)
```

## Set of random digits

```{r eval=FALSE}
classlist$course <- "2420"
classlist$course[grep("5242", classlist$Group)] <- "5242"
sample_class <- classlist %>% 
  separate(`Full name`, c("First", "Last"), sep=" ") %>%
  arrange(Last) %>% 
  select(First, Last, course) 
nbr <- data.frame(matrix(1:945, ncol=9, byrow=T))
sample_class <- cbind(sample_class, nbr)
write_csv(sample_class, "sample_classlist.csv")
```

```{r}
sample(0:9, 315, replace = TRUE)
```

## Just do it ...

True proportion of 2420:5242 is 0.59. Sample 20 students and check this. 

## Simpler approach computationally

- Use a sample function

```{r echo=TRUE, results='hide'}
sample_class %>% sample_n(20) 
```

##

```{r}
sample_class %>% sample_n(20) %>% kable()
```

## Statistical distributions

- Uniform
- Normal 
- Exponential
- Binomial
- Pareto
- Weibull
- Gamma
- Lognormal

## Random numbers = Uniform

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=sample(0:9, 10000, replace = TRUE), y=rep(0:9, 1000))
p1 <- ggplot(df, aes(x=y)) + geom_bar() + 
  scale_x_continuous("", breaks=0:10, labels=0:10) + ggtitle("Theory")
p2 <- ggplot(df, aes(x=x)) + geom_bar() + 
  scale_x_continuous("", breaks=0:10, labels=0:10) + ggtitle("Sample")
grid.arrange(p1, p2, ncol=2)
```

- symmetric, unimodal, uniform
- e.g. $U\{0, ..., 9\}$
- e.g. $P(X=x) = f(x) = 1/10, ~~ x \in \{0, ..., 9\}$

## Normal distribution

- Gaussian, bell-shaped
- symmetric, unimodal
- $N(\mu, \sigma)$

 $$f(x~|~\mu, \sigma) = \frac{1}{\sqrt{2}\pi\sigma} e^{-\frac{(x-\mu)^2}{2\sigma^2}}, ~~~ -\infty<x<\infty$$

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(-4,4, 0.008),  r=rnorm(1001), d=dnorm(seq(-4,4, 0.008)))
p1 <- ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + ggtitle("Theory")
p2 <- ggplot(df, aes(x=r, y=..density..)) + geom_histogram(binwidth=0.5) + xlab("") + ylab("") + geom_line(aes(x=x, y=d), color="red") + ggtitle("Sample")
grid.arrange(p1, p2, ncol=2)
```

## Exponential distribution

$$f(x~|~\lambda) = e^{-\lambda x} ~~ x\geq 0$$

- right-skewed, unimodal
- $Exp(\lambda)$
- Arises in time between or duration of events, e.g. time between successive failures of a machine, duration of a phone call to a help center

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(0,8, 0.008),  r=rexp(1001), d=dexp(seq(0,8, 0.008)))
p1 <- ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + ggtitle("Theory")
p2 <- ggplot(df, aes(x=r, y=..density..)) + geom_histogram(binwidth=0.5) + xlab("") + ylab("") + geom_line(aes(x=x, y=d), color="red") + ggtitle("Sample")
grid.arrange(p1, p2, ncol=2)
```

## Poisson distribution

$$P(X=x~|~\lambda) = \frac{\lambda^x e^{-\lambda}}{x!} ~~ x \in \{0, 1, 2, ...\}$$

- discrete, right-skewed, unimodal
- Arises when counting number of times event occurs in an interval of time, e.g. the number of patients arriving in an emergency room between 11 and 12 pm

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(0, 8, 1),  d=dpois(seq(0, 8, 1), lambda=1))
p1 <- ggplot(df, aes(x=x, y=d)) + 
  geom_bar(stat="identity") + xlab("") + ylab("") + xlim(c(-1,8)) +
  geom_hline(yintercept=0, colour="grey70") + ggtitle("Theory")
df2 <- data.frame(r=rpois(1001, lambda=1))
p2 <- ggplot(df2, aes(x=r)) + geom_bar() +
  xlab("") + ylab("") + xlim(c(-1,8)) + ggtitle("Sample")
grid.arrange(p1, p2, ncol=2)
```

## Binomial

$$P(X=x~|~n,p) = \left(\begin{array}{c} n \\ p \end{array} \right) p^x (1-p)^{n-x} ~~ x \in \{0, 1, 2, ..., n\}$$

- discrete, unimodal, right- or left-skewed or unimodal depending on $p$
- Arises from counting the number of successes from $n$ independent Bernouilli trials, e.g. the number of heads in 10 coin flips

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(0, 10, 1),  d=dbinom(seq(0, 10, 1), 10, 0.5))
p1 <- ggplot(df, aes(x=x, y=d)) + 
  geom_bar(stat="identity") + xlab("") + ylab("") + 
  scale_x_continuous(breaks=seq(0,10,1), labels=seq(0,10,1)) + 
  ggtitle("Theory")
df2 <- data.frame(r=rbinom(100, 10, 0.5))
p2 <- ggplot(df2, aes(x=r)) + geom_bar() +
  scale_x_continuous(breaks=seq(0,10,1), labels=seq(0,10,1), limits=c(0,10)) + 
  xlab("") + ylab("") + ggtitle("Sample")
grid.arrange(p1, p2, ncol=2)
```

## Pareto

```{r}
dpareto <- function(x, c) {
  # x must be greater than 1
  if (c<=0) stop("c must be positive") # Diagnostic step
  c/x^(c+1)
}

ppareto <- function(q, c) {
  if (c<=0) stop("c must be positive > 0")
  ifelse(q<1, 0, 1-1/q^c)
}

qpareto <- function(p,c){
  if (c<=0) stop("c must be positive > 0")
  if (any(p<0)|any(p>1)) # Symbol | denotes logical OR
    stop("p must be between 0 and 1")
  q <- (1-p)^(-1/c)
  q
}

rpareto <- function(n, c){
 if (c<=0) stop("c must be positive")
   rp <- runif(n)^(-1/c)
  rp
}
```

$$f(x~|~\alpha, \lambda) = \frac{\alpha\lambda^\alpha}{(\lambda+x)^{\alpha+1}} ~~~x>0, \alpha>0, \lambda>0 $$ 

- Used to describe allocation of wealth, sizes of human settlement
- Heavier tailed than exponential distribution

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(1, 10, 0.1),  d=dpareto(seq(1, 10, 0.1), 1))
p1 <- ggplot(df, aes(x=x, y=d)) + 
  geom_line() + xlab("") + ylab("") +
  scale_x_continuous(breaks=seq(1,10,1), labels=seq(1,10,1)) +
  ggtitle("Theory")
df2 <- data.frame(r=rpareto(1000, 1))
p2 <- ggplot(df2, aes(x=r)) + geom_histogram(binwidth=1) +
  scale_x_continuous(breaks=seq(1,10,1), labels=seq(1,10,1), limits=c(0.5,10.5)) +
  xlab("") + ylab("") + ggtitle("Sample")
grid.arrange(p1, p2, ncol=2)
```


## Weibull

$$f(x~|~\lambda, k) = \frac{k}{\lambda}\left(\frac{x}{\lambda}\right)^{k-1} e^{(-x/\lambda)^k}, ~~~ x\geq 0$$

- used for particle size distribution, failure analysis, delivery time, extreme value theory
- shape changes considerably with different $k$

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(0.0025, 2.5, 0.0025),  r=rweibull(1000, shape=1.5), d=dweibull(seq(0.0025, 2.5, 0.0025), shape=1.5))
p1 <- ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + ggtitle("Theory") + xlim(c(0,4.5))
p2 <- ggplot(df, aes(x=r, y=..density..)) + geom_histogram(binwidth=0.5) + xlab("") + ylab("") + geom_line(aes(x=x, y=d), color="red") + 
  ggtitle("Sample") + xlim(c(0,4.5))
grid.arrange(p1, p2, ncol=2)
```

## Gamma

$$f(x~|~\alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1}e^{-x\beta}, ~~~ x\geq 0 ~~\alpha, \beta > 0 $$

- Generalisation of exponential distribution, and also $\chi^2$
- $\alpha$ changes shape substantially
- used to model size of insurance claims, rainfall

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(0.005, 5, 0.005),  r=rgamma(1000, shape=2), d=dgamma(seq(0.005, 5, 0.005), shape=2))
p1 <- ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + ggtitle("Theory") + xlim(c(0,4.5))
p2 <- ggplot(df, aes(x=r, y=..density..)) + geom_histogram(binwidth=0.5) + xlab("") + ylab("") + geom_line(aes(x=x, y=d), color="red") + 
  ggtitle("Sample") + xlim(c(0,4.5))
grid.arrange(p1, p2, ncol=2)
```

## Lognormal

- Also called Galton's distribution
- Generated when $Y\sim N(\mu, \sigma)$, and study $X=exp(Y)$
- used for modeling length of comments posted in internet discussion forums, users' dwell time on the online articles, size of living tissue, highly communicable epidemics

```{r fig.align='center', fig.width=8, fig.height=3, fig.show='hold'}
df <- data.frame(x=seq(0.005, 5, 0.005),  r=rlnorm(1000), d=dlnorm(seq(0.005, 5, 0.005)))
p1 <- ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + ggtitle("Theory") + xlim(c(0,4.5))
p2 <- ggplot(df, aes(x=r, y=..density..)) + geom_histogram(binwidth=0.5) + xlab("") + ylab("") + geom_line(aes(x=x, y=d), color="red") + 
  ggtitle("Sample") + xlim(c(0,4.5))
grid.arrange(p1, p2, ncol=2)
```

## Sampling variability

```{r fig.align='center', fig.width=8, fig.height=5}
df <- data.frame(x=sample(0:9, 50, replace = TRUE))
p1 <- ggplot(df, aes(x=x)) + geom_bar() + 
  scale_x_continuous("", breaks=0:10, labels=0:10) + ggtitle("n=50")
df <- data.frame(x=sample(0:9, 100, replace = TRUE))
p2 <- ggplot(df, aes(x=x)) + geom_bar() + 
  scale_x_continuous("", breaks=0:10, labels=0:10) + ggtitle("n=100")
df <- data.frame(x=sample(0:9, 1000, replace = TRUE))
p3 <- ggplot(df, aes(x=x)) + geom_bar() + 
  scale_x_continuous("", breaks=0:10, labels=0:10) + ggtitle("n=1000")
df <- data.frame(x=sample(0:9, 100000, replace = TRUE))
p4 <- ggplot(df, aes(x=x)) + geom_bar() + 
  scale_x_continuous("", breaks=0:10, labels=0:10) + ggtitle("n=100000")
grid.arrange(p1, p2, p3, p4, ncol=2)
```

## Probability calculations

- Probability density functions are useful for computing expected quantities
- E.g. Gamma(2,1), what is the probability of seeing $X>3.2$, or $1.5<X<2.5$

```{r echo=TRUE}
pgamma(3.2, 2, lower.tail=FALSE)
pgamma(2.5, 2) - pgamma(1.5, 2)
```

```{r fig.align='center', fig.width=7, fig.height=2.5, fig.show='hold'}
df <- data.frame(x=seq(0.005, 5, 0.005), d=dgamma(seq(0.005, 5, 0.005), shape=2))
p1 <- ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + 
  geom_vline(xintercept=3.2, colour="red") +
  ggtitle("X>3.2") + xlim(c(0,4.5))
p2 <- ggplot(df, aes(x=x, y=d)) + geom_line() + xlab("") + ylab("") +
  geom_hline(yintercept=0, colour="grey70") + 
  geom_vline(xintercept=c(1.5, 2.5), colour="red") +
  ggtitle("1.5<X<2.5") + xlim(c(0,4.5))
grid.arrange(p1, p2, ncol=2)
```

## Your turn

- Continuous distributions: Area under the curve = ______
- Discrete distributions: Sum of probabilities = ______

## Resources

- [NIST Statistics Handbook](http://www.itl.nist.gov/div898/handbook/eda/section3/eda366.htm)
- [random.org](https://www.random.org/randomness/)
- [Radioactive decay](https://www.fourmilab.ch/hotbits/)
- [electromagnetic field of a vacuum](https://qrng.anu.edu.au)
- [wikipedia](https://en.wikipedia.org/wiki/List_of_probability_distributions)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
