---
title: "Fitting and Diagnosing <br/> Multiple Linear Regression Models"
author: "Dilini Talagala"
output:
  xaringan::moon_reader:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  echo = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 8,
  fig.width = 10,
  fig.align = "center",
  cache = FALSE
)
library(knitr)
```

###  Let's take a look at the OECD PISA 2015 data focusing on Australia

```{r echo=TRUE}
library("tidyverse")
library("forcats")
load("pisa_au.rda")
```

#### About the dataset

- There are ten values for each student for the science score. 
- The raw scores that a student earns in the test are not distributed.
- But rather a large linear model is constructed, and ten predictions are randomly generated for each student from the model.

---
### Scatterplot matrix of the plausible scores for each student of Australia


```{r echo=TRUE, eval=FALSE}
library("GGally")

sci_scores <- pisa_au %>% select(PV1SCIE, PV2SCIE, PV3SCIE,
                                 PV4SCIE, PV5SCIE, PV6SCIE,
                                 PV7SCIE, PV8SCIE, PV9SCIE,
                                 PV10SCIE) %>% as.data.frame()

ggscatmat(sci_scores, alpha=0.1)
```
---
```{r eval=TRUE,  fig.height=8}
library("GGally")
sci_scores <- pisa_au %>% select(PV1SCIE, PV2SCIE, PV3SCIE,
                                 PV4SCIE, PV5SCIE, PV6SCIE,
                                 PV7SCIE, PV8SCIE, PV9SCIE,
                                 PV10SCIE) %>% as.data.frame()

ggscatmat(sci_scores, alpha=0.1)
```

---

### Create a new variable which is the average of the ten scores for each student

```{r echo=TRUE}
pisa_au <- pisa_au %>% 
  mutate(science = (PV1SCIE + PV2SCIE + PV3SCIE + PV4SCIE + 
                      PV5SCIE + PV6SCIE + PV7SCIE + PV8SCIE +
                      PV9SCIE + PV10SCIE) / 10)
```
---
### Students are tested at many different schools. How many schools?

```{r echo=TRUE}
pisa_au %>% group_by(CNTSCHID) %>% tally() %>% 
  arrange(desc(n)) -> aus_schools

dim(aus_schools)

```

```{r echo = TRUE}
head(aus_schools)
```

---
### Distribution of number of students tested at each school


```{r echo=TRUE, fig.height= 6}
ggplot(aus_schools, aes(x = n)) + 
  geom_density(fill = "black", alpha = 0.5)
```

---

### Selecting a subset of data  

```{r echo=TRUE}
pisa_au <- pisa_au %>% 
  select(science, ST004D01T, OUTHOURS, ANXTEST, EMOSUPP,
         PARED, JOYSCIE, WEALTH, ST013Q01TA, ST012Q01TA,
         SENWT)

glimpse(pisa_au)
```
---
##### Make summaries for each of the variables, to examine their suitability for modeling.
```{r echo=TRUE, size="large"}
summary(pisa_au)
```

---

class: inverse, center, middle

# Handling missing data


---
#### EMOSUPP is all missing, cannot use this variable.

Actions to take:
- Drop EMOSUPP

```{r echo=TRUE}
pisa_au <- pisa_au %>% select(-EMOSUPP)

glimpse(pisa_au)

```

---

- **OUTHOURS has about one third missing, might be unreliable**
- **JOYSCIE has 10% missing, might be unreliable**

Actions to take:
- Remove any case with missing values and check numbers remaining


```{r echo=TRUE}
aus_nomiss <- pisa_au %>% filter(!is.na(OUTHOURS)) %>%
  filter(!is.na(ANXTEST)) %>% filter(!is.na(PARED)) %>%
  filter(!is.na(JOYSCIE)) %>% filter(!is.na(WEALTH)) %>%
  filter(!is.na(ST013Q01TA)) %>% filter(!is.na(ST012Q01TA))

nrow(pisa_au)

nrow(aus_nomiss)

```

---

### Let's look at the distribution of weights

- The students all have weights associated with them. 
- This is an indication of how many other students they represent in Australia, relative to their socioeconomic and demographic characteristics. 


```{r fig.height= 4, echo=TRUE}
ggplot(aus_nomiss, aes(x=SENWT)) + 
  geom_density(fill = "black", alpha = 0.5)
```
---

#### The weights are bimodal with a few very large ones

- Is the bimodality due to one of the variables in the study that we are using for the model? *It is not due to gender!*


```{r fig.height = 5, echo = TRUE}
ggplot(aus_nomiss, aes(x=SENWT)) + 
  facet_wrap(~ST004D01T) +
  geom_density(fill="black", alpha=0.5)
```
---

class: inverse, center, middle

# Model building 

Response: `science` (standardised)

Explanatory variables:  all the remaining variables


---
### Transforming data

- Some variables need to be treated as categorical variables
- So it is best if they are forced to be factors before modeling:


```{r echo=TRUE}
aus_nomiss <- aus_nomiss %>%
  mutate(ST004D01T = factor(ST004D01T, 
                            levels=c(1,2),
                            labels=c("f", "m")))
```

- Standardising the response variable: `science` 
```{r echo=TRUE, out.height= 3}
aus_nomiss <- aus_nomiss %>% 
  mutate(science_std = (science - mean(science)) / sd(science))

```

---

##### Fitting weighted multiple regression model for science against gender, and joy of science. 

```{r echo=TRUE, size="small"}
aus_glm_test <- glm(science_std ~ ST004D01T + JOYSCIE,
                    data = aus_nomiss, weights = SENWT)

summary(aus_glm_test)
```

---
### Sketch what this model looks like.

```{r echo=TRUE, fig.height= 4}
coefs <- coefficients(aus_glm_test)

ggplot() + 
  geom_abline(intercept=coefs[1], slope=coefs[3], colour="red")+
  geom_abline(intercept=coefs[1]+coefs[2], slope=coefs[3],
              colour="blue") +
  xlim(range(aus_nomiss$JOYSCIE)) + ylim(c(-1,1))

```

---

Make plots of the response variable `science_std` against each of the possible explanatory variables. 

A good selection of plots are:

- side-by-side boxplots for the categorical (or discrete) predictors

- a smoother for the continuous predictors.

---


```{r echo = TRUE, eval = TRUE, fig.height=3}
library(gridExtra)

p1 <- ggplot(aus_nomiss, aes(x=ST004D01T, y=science_std)) + 
  geom_boxplot() + xlab("Gender")
p2 <- ggplot(aus_nomiss, aes(x=factor(PARED), y=science_std)) +
  geom_boxplot() + xlab("Parents years of schools")
p3 <- ggplot(aus_nomiss, aes(x=OUTHOURS, y=science_std)) +
  geom_smooth() + xlab("Out-of-School Study Time")
p4 <- ggplot(aus_nomiss, aes(x=WEALTH, y=science_std)) +
  geom_smooth() + xlab("Family wealth (WLE)")

grid.arrange(p1, p2, p3, p4, ncol=4)
```
---
## Your turn

Add the following variables to the same grid

- ST012Q01TA: ("How many Televisions are there in your home ") 
- ST013Q01TA: ("How many books are there in your home")
- ANXTEST: ("Personality: Test Anxiety (WLE)")
- JOYSCIE: ("Enjoyment of science (WLE)")

---
### Compute the leverage and influence statistics

```{r echo = TRUE, fig.height= 5, eval = FALSE}
library("broom")
model_augment <- augment(your-model-here)

ggplot(model_augment, aes(x=.hat, y=.cooksd)) +
  geom_point()+
  theme(aspect.ratio=1)
```
---

### Compute the variance inflation factors

```{r echo= TRUE, eval = FALSE}
library("car")

vif(your-model-here)
```

---

class: inverse, center, middle


## Resources

[Statistics online textbook, Diez, Barr, Cetinkaya-Rundel](https://www.openintro.org/stat/textbook.php?stat_book=isrs). 


# Happy learning with R :)