---
title: "ETC 2420/5242 Lab 1 2016"
author: "Di Cook"
date: "SOLUTION"
output: pdf_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```


## Question 1 (2pts)

There could be many possible questions that might be answered by this data. Examples include these ones:

- Does the personal savings rate dip when unemployment is high?

- Is there a seasonal effect in unemployment?

- Is population increasing?

## Question 2 (2pts)

There could be many possible questions that might be answered by this data. Examples include these ones:

- Is life expectancy positively associated with gdp percapita?

- Is life expectancy increasing over time?

- Is the trend in life expectancy similar across all countries?

## Question 3 (2pts)

There could be many possible questions that might be answered by this data. Examples include these ones:

- What places in the city see the most pedestrians?

- What times would be rush hours on week days?

- Can you see the Wednesday night markets location and time based on pedestrian traffic?

- Is White Night visible in terms of pedestrian traffic?

- Are more people out and about in summer than in winter?


## Question 4 (4pts)

1. Read in the OECD PISA data
2. (1pt) Tabulate the countries (CNT)
3. Extract the values for Australia (AUS) and Shanghai (QCN)
4. (1pt) Compute the average and standard deviation of the reading scores (PV1READ), for each country
5. (2pts) Write a few sentences explaining what you learn about reading scores in these two countries.

```{r}
student2012.sub <- readRDS("../data/student_sub.rds") 
table(student2012.sub$CNT)
australia <- student2012.sub[student2012.sub$CNT=="AUS",]
shanghai <- student2012.sub[student2012.sub$CNT=="QCN",]
mean(australia$PV1READ)
sd(australia$PV1READ)
mean(shanghai$PV1READ)
sd(shanghai$PV1READ)
```

The reading scores are higher in Shanghai than in Australia by about 67 points. The variation in scores in Australia is higher, with a standard deviation of 100 as opposed to 80 for Shanghai.

```{r}
# Alternative way to do the code
library(dplyr)
library(knitr)
library(tidyr)
student2012.sub %>% select(CNT) %>% group_by(CNT) %>% tally() 
student2012.sub %>% filter(CNT %in% c("AUS", "QCN")) %>%
  group_by(CNT) %>%
  summarise(m=mean(PV1READ), s=sd(PV1READ)) %>% kable(digits=1)
```
