---
title: "ETC 2420/5242 Lab 4 2016"
author: "Di Cook"
date: "Week 4"
output: pdf_document
---

```{r echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  echo = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

## Purpose

This lab is to examine different statistical distributions, fit distributions to samples by estimating the parameters by maximum likelihood and checking the fit with QQ-plots. 

## Reading

Read the material on maximum likelihood estimation at [https://onlinecourses.science.psu.edu/stat414/node/191](https://onlinecourses.science.psu.edu/stat414/node/191). 

Read the code in the lecture notes from Week 3. Particularly look at the functions for making QQ-plots, computing and plotting the likelihood functions.

## Warmup exercises

- Compute these probabilities for $X\sim N(2, 0.5)$
    + $P(X<1.3)$
    + $P(X>1.9)$
    + $P(1.8<X<2.2)$
- Compute the quantile value (X) for $X\sim N(-3, 4)$ which matches these probabilities
    + $P(X<x)=0.53$
    + $P(X<x)=0.12$
    + $P(X<x)=0.84$
    + $P(X<x)=1.2$
- Compute the value of the density function for a $N(-1.2, 0.8)$ corresponding to $X=$
    + -1.0
    + -0.2
    + -2.0
- Plot the density curves a 
    + $Weibull(2.5, 2)$
    + $Weibull(2.5, 1)$
    + $Weibull(1, 1)$
on the same plot.

## Question 1

1. Simulate samples of size $n=$ 30, 100, 500 from these distributions
    a. Lognormal(2, 0.5)
    b. Gamma(2, 4)
    
2. Make a QQ-plot of each these samples. 

## Question 2

Using this code, generate a sample of size $n=267$ from a $Gamma(1.2, 0.25)$ distribution.

```{r echo=TRUE}
set.seed(123)
X2 <- data.frame(x=rgamma(n=267, 1.2, 0.25))
```

a. Plot the sample, using a histogram, describe the shape of the distribution.
b. What parameters of the gamma distribution were used to simulate the sample? ($\alpha, \beta$)
c. If we are to use maximum likelihood distribution what values would we expect to get as the parameter estimates?
d. Write a function to compute the likelihood function. 
e. Plot the likelihood function for a range of values of $\alpha, \beta$ that shows the maximum likelihood estimates for each parameter.
f. Look up the function `fitdistr` from the `MASS` library. Explain what this does. Use it to find the MLE estimates for $\alpha, \beta$. How do these compare with the values you read off your plot?

## Question 3

Take a look at the data set `danishuni` from the `CASdatasets` library. 

```{r echo=TRUE}
library(CASdatasets)
data(danishuni)
```

a. Make a histogram of the `Loss`. Describe the shape.
b. Fit both a gamma and lognormal distribution to the sample, i.e. find the MLEs.  
c. Produce a QQ-plot for each of the distributions. 
d. Which is the better fit to the sample?
e. Re-do steps a-d after log-transforming the `Loss`.

## TURN IN 

- Your `.Rmd` file
- Your Word (or pdf) file that results from knitting the Rmd.
- Make sure your group members are listed as authors, one person per group will turn in the report
- DUE: Wednesday after the lab, by 7am, loaded into moodle



## Resources


- [PSU lecture notes on MLE](https://onlinecourses.science.psu.edu/stat414/node/191)
- [CASdatasets](http://cas.uqam.ca)


