---
title: "ETC 2420/5242 Lab 8 2016"
author: "Di Cook"
date: "Week 8"
output: pdf_document
---

```{r echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  echo = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
library(knitr)
```

## Data

```{r}
# Read stations data
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggmap)
stations <- read_table("../data/ghcnd-stations.txt", 
  col_names=c("ID", "lat", "lon", "elev", "state", "name", 
              "v1", "v2", "v3"), skip=353, n_max=17081)

oz <- map_data("world", xlim=range(stations$lon),
               ylim=range(stations$lat))
ggplot(oz, aes(x=long, y=lat)) + geom_path(aes(group=group)) + 
  coord_quickmap() + 
  geom_point(data=stations, aes(x=lon, y=lat), 
             colour="red", alpha=0.5)

# Get pedestrian sensor locations
ped_loc <- read_csv("../data/Pedestrian_Sensor_Locations.csv")

melb <- get_map(location=c(mean(range(ped_loc$Longitude)),
                           mean(range(ped_loc$Latitude))), zoom=14)
ggmap(melb) + geom_point(data=ped_loc, 
                         aes(x=Longitude, y=Latitude), 
                         colour="#c51b7d", alpha=0.5, size=3)

# Choose a weather station close to pedestrian sensors
melb_stns <- stations %>% filter(lon > min(ped_loc$Longitude), 
                                 lon < max(ped_loc$Longitude), 
                                 lat > min(ped_loc$Latitude),
                                 lat < max(ped_loc$Latitude))

ggmap(melb) + geom_point(data=ped_loc, 
                         aes(x=Longitude, y=Latitude), 
                         colour="#c51b7d", alpha=0.5, size=3) +
  geom_point(data=melb_stns, aes(x=lon, y=lat), 
             colour="#542788", size=6, shape=18)

# Read temp data
melb_ghcn <- read_csv("../data/melb_ghcn.csv")
melb_ghcn <- melb_ghcn %>% 
  select(stn_id, date, variable, value) %>%
  filter(variable %in% c("TMAX", "TMIN", "PRCP")) %>%
  mutate(year=substr(date, 1, 4), month=substr(date, 5, 6),
         mdate=substr(date, 7, 8)) %>%
  mutate(date=as.Date(paste(mdate, month, year, sep="-"),
                      format="%d-%m-%Y")) %>%
  mutate(mdate=as.numeric(mdate), month=as.numeric(month), 
         year=as.numeric(year), value = value/10)
melb_ghcn_wide <- melb_ghcn %>% spread(variable, value)

# Read sensor counts
ped_sub <- read_csv("../data/pedestrian_counts_sub.csv")

# 

```

## Reading

- Read the code in the lecture notes on computing bootstrap confidence intervals for linear models from Week 6. 

```{r}
library("tidyr")
library("dplyr")
library("ggplot2")
student2012.sub <- readRDS("../data/student_sub.rds") 
australia <- student2012.sub[student2012.sub$CNT=="AUS",]
australia <- australia %>% mutate(math = (PV1MATH+PV2MATH+PV3MATH+PV4MATH+PV5MATH)/5)
```

The variables that were used for modeling `math` was:

\begin{tabular}{llp{3in}}
Variable name & Description & Coding\\\hline
ST04Q01 & Gender & 1=Female, 2=Male\\
ST06Q01 & Age when started school & Actual age, 9997-9999 indicate missing values\\
ST15Q01 & Mother Current Job Status & 1=Full-time, 2=Part-time, 3=Not working, but looking for a job, 4=Other (inc stay-at-home), 7-9 indicate missing values\\
ST19Q01 & Father Current Job Status &  1=Full-time, 2=Part-time, 3=Not working, but looking for a job, 4=Other (inc stay-at-home), 7-9 indicate missing values\\
ST26Q01 & Possessions - desk  & 1=Yes, 2=No, 7-9 indicate missing values\\
ST26Q04 & Possessions - computer  & 1=Yes, 2=No, 7-9 indicate missing values\\
ST26Q06 & Possessions - Internet  & 1=Yes, 2=No, 7-9 indicate missing values\\
ST27Q02 & How many - televisions & 1=None, 2=One, 3=Two, 4=Three or more,  7-9 indicate missing values\\
ST28Q01 & How many books at home & 1=0-10, 2=11-25, 3=26-100, 4=101-200, 5=201-500, 6=More than 500, 7-9 indicate missing values\\
SENWGT\_STU & Weight & Reflects how the student represents other students in Australia based on socioeconomic and demographic characteristics
\end{tabular}

```{r}
australia <- australia %>% 
  select(math, SCHOOLID, ST04Q01, ST06Q01, ST57Q01, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q04, ST26Q06, ST27Q02, ST28Q01, SENWGT_STU)
australia$ST06Q01[australia$ST06Q01 > 9990] <- NA
australia$ST57Q01[australia$ST57Q02 > 9990] <- NA
australia$ST15Q01[australia$ST15Q01 > 6] <- NA
australia$ST19Q01[australia$ST19Q01 > 6] <- NA
australia$ST26Q01[australia$ST26Q01 > 6] <- NA
australia$ST26Q02[australia$ST26Q02 > 6] <- NA
australia$ST26Q04[australia$ST26Q04 > 6] <- NA
australia$ST26Q06[australia$ST26Q06 > 6] <- NA
australia$ST27Q02[australia$ST27Q02 > 6] <- NA
australia$ST28Q01[australia$ST28Q01 > 6] <- NA
australia <- australia %>% select(-ST57Q01)
aus_nomiss <- australia %>% filter(!is.na(ST04Q01)) %>%
  filter(!is.na(ST06Q01)) %>% filter(!is.na(ST15Q01)) %>%
  filter(!is.na(ST19Q01)) %>% filter(!is.na(ST26Q01)) %>%
  filter(!is.na(ST26Q02)) %>% filter(!is.na(ST26Q04)) %>%
  filter(!is.na(ST26Q06)) %>% filter(!is.na(ST27Q02)) %>%
  filter(!is.na(ST28Q01))
```


Model building will be done using:

- Response: `math` (standardised)
- Explanatory variables: `ST04Q01, ST06Q01, ST15Q01, ST19Q01, ST26Q01,  ST26Q04, ST26Q06, ST27Q02, ST28Q01`.

```{r}
aus_nomiss$SCHOOLID <- factor(aus_nomiss$SCHOOLID)
aus_nomiss$ST04Q01 <- factor(aus_nomiss$ST04Q01)
aus_nomiss$ST15Q01 <- factor(aus_nomiss$ST15Q01)
aus_nomiss$ST15Q01 <- factor(aus_nomiss$ST15Q01)
aus_nomiss$ST19Q01 <- factor(aus_nomiss$ST19Q01)
aus_nomiss$ST26Q01 <- factor(aus_nomiss$ST26Q01)
aus_nomiss$ST26Q02 <- factor(aus_nomiss$ST26Q02)
aus_nomiss$ST26Q04 <- factor(aus_nomiss$ST26Q04)
aus_nomiss$ST26Q06 <- factor(aus_nomiss$ST26Q06)
aus_nomiss <- aus_nomiss %>% mutate(math_std = (math-mean(math))/sd(math))
aus_nomiss$ST06Q01 <- aus_nomiss$ST06Q01 - 4
```

## Question 1

```{r}
aus_glm <- glm(math_std~ST04Q01+ST06Q01+ST15Q01+ST19Q01+ST26Q01+ST26Q04+ST26Q06+ST27Q02+ST28Q01, data=aus_nomiss, weights=SENWGT_STU)
```

a. Compute and report the 95% confidence interval for the parameter for the number of books in the household (ST28Q01), using classical t-interval methods.
b. Use this to test the hypothesis that ST28Q01 is not important for the model.

## Question 2

a. The boot package can generate bootstrap samples for weighted data. To use the `boot` function for drawing samples, you need a function to compute the statistic of interest. Write the function to return the slope for ST28Q01 after fitting a `glm` to a bootstrap sample. The skeleton of the function `calc_stat` is below, where `d` is the data, and `i` is the vector of indices of the bootstrap sample. 

```{r eval=FALSE, echo=TRUE}
library(boot)
calc_stat <- function(d, i) {
  x <- d[i,]
  mod <- FILL IN THE NECESSARY CODE
  stat <- FILL IN THE NECESSARY CODE
  return(stat)
}
stat <- boot(aus_nomiss, statistic=calc_stat, R=1000,
     weights=aus_nomiss$SENWGT_STU)
stat
sort(stat$t)[25]
sort(stat$t)[975]
```

b. How does the bootstrap interval compare with the t-interval?

## Question 3

Now make a 95% bootstrap confidence interval for predicted value for a new student who is FEMALE, started school at 4, mother and father both work full-time, has a desk, computer and internet, two TVs and 26-100 books in the home. The weight for a student like this is 0.1041. Be sure to convert the values back into the actual math score range.

## Question 4

Compute a bootstrap 95% prediction interval for the same student as in the previous question. Be sure to convert the values back into the actual math score range.

## TURN IN 

- Your `.Rmd` file
- Your Word (or pdf) file that results from knitting the Rmd.
- Make sure your group members are listed as authors, one person per group will turn in the report
- DUE: Wednesday after the lab, by 7am, loaded into moodle



## Resources

- https://data.melbourne.vic.gov.au/Transport-Movement/Pedestrian-Sensor-Locations/ygaw-6rzq
- Menne, M.J., I. Durre, B. Korzeniewski, S. McNeal, K. Thomas, X. Yin, S. Anthony, R. Ray, R.S. Vose, B.E.Gleason, and T.G. Houston, 2012: Global Historical Climatology Network - Daily (GHCN-Daily), Version 3. [indicate subset used following decimal, 
e.g. Version 3.12]. NOAA National Climatic Data Center. http://doi.org/10.7289/V5D21VHZ [9/9/2016].




