---
title: "ETC 2420/5242 Lab 9 2016"
author: "Di Cook"
date: "SOLUTION"
output: pdf_document
---

```{r echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  echo = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 3,
  fig.width = 3,
  fig.align = "center",
  cache = FALSE
)
library(knitr)
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggmap)
library(lubridate)
library(broom)
```


```{r fig.width=8, fig.height=8, fig.align='center'}
load("../data/ped_weather_sub_glm.rda")
ped_weather_sub <- ped_weath_sub_glm$data
```


## Question 1

Write down the equation of the model that was used to play the game.

$$ \hat{log{(y_i})} = \beta_0 + \beta_1 Tues + \beta_2Wed +...+\beta_6Sun + $$
$$\beta_7time_2 + ... + \beta_{29}time_{24} +$$
$$\beta_{30}Feb + ... + \beta_{41}Dec + \beta_{41}MelbCent+ $$
$$<two-way, ~three-way ~and~ four-way ~interaction ~terms> +$$ $$\beta_{4033}high tmp not+\beta_{4034}low tmp not+\beta_{4035}high prcp none$$

## Question 2

a. Simulate 10 sets of new values using the `simulate` function.

```{r}
ped_new <- simulate(ped_weath_sub_glm, 10)
ped_new <- bind_cols(ped_weather_sub, ped_new)
```

b. Subset to examine only records for March 28, 9am. (What day of the week is this?)

```{r}
ped_new_sub <- ped_new %>% 
  filter(date == ymd("2013-03-28"), time==9) %>%
  select(sensor_name, count, sim_1:sim_10) %>%
  gather(sim, value, sim_1:sim_10)
wday(ymd("2013-03-28"), label=TRUE)
```

c. Summarise the distribution of the 10 values for the two locations, and compare with the actual count.

```{r fig.width=6, fig.align='center'}
ggplot(ped_new_sub, aes(x=sensor_name, y=value)) + geom_point() +
  geom_point(aes(y=count), colour="red", size=3) + 
  xlab("") + ylab("Count")
```

d. If you are open, have three attendants at Flinders and two attendants at Melbourne Central, how much would you make at most, and at least under these conditions, at each location?

```{r}
sample_day_time <- function(){
  sel_date <- sample(ped_weather_sub$date, 1)
  sel_time <- sample(7:21, 1)
  return(list(sel_date, sel_time))
}
dt <- sample_day_time()
compute_earnings <- function(sel_date, sel_time, Fl_attendants=1, MC_attendants=1) {
  Fl_rate <- c(rep(0.1, 3), rep(0.05, 6), rep(0.01, 4), rep(0, 2))[sel_time-6]
  MC_rate <- c(rep(0.08, 3), rep(0.06, 6), rep(0.02, 6))[sel_time-6]
  ped_sim <- ped_weather_sub
  ped_sim$new1 <- simulate(ped_weath_sub_glm)$sim_1
  ped_sim_sub <- ped_sim %>% filter(date == sel_date, time == sel_time)
  Fl_count <- ped_sim_sub %>% 
    filter(sensor_name == "Flinders Street Station Underpass") %>% 
    select(new1)
  MC_count <- ped_sim_sub %>% 
    filter(sensor_name == "Melbourne Central") %>% 
    select(new1)
  Fl <- 0
  MC <- 0
  if (nrow(Fl_count) > 0) Fl <- min(round(Fl_count*Fl_rate, 0), 50*Fl_attendants)
  if (nrow(MC_count) > 0) MC <- min(round(MC_count*MC_rate, 0), 50*MC_attendants)
  Fl_earn <- 0
  MC_earn <- 0
  if (Fl > 0) Fl_earn <- Fl*4 - (100+50*(Fl_attendants-1))
  if (MC > 0) MC_earn <- MC*4 - (100+50*(MC_attendants-1))
  return(list(Fl_earn, MC_earn))
}
sel_date <- ymd("2013-03-28")
sel_time <- 9
compute_earnings(sel_date, sel_time, 3, 2)
```

```{r eval=FALSE}
Flinders it would be 1262*.1*4-200=304.8 at highest and 1150*.1*4-200=260 at the lowest. Melbourne Central it would be 1074*.08*4-150=193.68 at the highest, and 957*.08*4-150=156.24 at the lowest.
```

## Question 3

a. Now extend this to the full day (between 7am-10pm, closing at 10pm), keep the same number of attendants for the full day. How much in profit do you make at most, and at the lowest at each location?

```{r}
earn_Fl <- NULL
earn_MC <- NULL
for (i in 7:21) {
  earn <- compute_earnings(sel_date, i, 3, 2)
  earn_Fl <- c(earn_Fl, earn[[1]])
  earn_MC <- c(earn_MC, earn[[2]])
}
earn_Fl
earn_MC
sum(earn_Fl)
sum(earn_MC)
```

b. Suppose that the weather for the day is actually a hot day. How does this affect your profits? (Predict the counts for both locations for the hot and not hot day, and compute the difference. Subtract this number from your simulation values - because these were for a not hot day.)

```{r}
newdat <- data.frame(day=c("Thu", "Thu", "Thu", "Thu"), 
                     month=c("Mar", "Mar", "Mar", "Mar"), 
                     time=c(9, 9, 9, 9), 
                     sensor_name=c("Melbourne Central", "Melbourne Central",
                                   "Flinders Street Station Underpass", 
                                   "Flinders Street Station Underpass"),
                     high_tmp=c("hot","not", "hot","not"), 
                     low_tmp=c("not", "not", "not", "not"), 
                     high_prcp=c("none", "none", "none", "none"))
newdat$time <- factor(newdat$time, levels=0:23)
pred <- exp(predict(ped_weath_sub_glm, newdat))
Fl_dif <- pred[4] - pred[3]
MC_dif <- pred[2] - pred[1]
```

```We need to modify the compute_earnings function to take the reduction in pedestrians based on the weather into account.```

```{r}
compute_earnings2 <- function(sel_date, sel_time, Fl_attendants=1, MC_attendants=1) {
  Fl_rate <- c(rep(0.1, 3), rep(0.05, 6), rep(0.01, 4), rep(0, 2))[sel_time-6]
  MC_rate <- c(rep(0.08, 3), rep(0.06, 6), rep(0.02, 6))[sel_time-6]
  ped_sim <- ped_weather_sub
  ped_sim$new1 <- simulate(ped_weath_sub_glm)$sim_1
  ped_sim_sub <- ped_sim %>% filter(date == sel_date, time == sel_time)
  Fl_count <- ped_sim_sub %>% 
    filter(sensor_name == "Flinders Street Station Underpass") %>% 
    select(new1) - Fl_dif
  MC_count <- ped_sim_sub %>% 
    filter(sensor_name == "Melbourne Central") %>% 
    select(new1) - MC_dif
  Fl <- 0
  MC <- 0
  if (nrow(Fl_count) > 0) Fl <- min(round(Fl_count*Fl_rate, 0), 50*Fl_attendants)
  if (nrow(MC_count) > 0) MC <- min(round(MC_count*MC_rate, 0), 50*MC_attendants)
  Fl_earn <- 0
  MC_earn <- 0
  if (Fl > 0) Fl_earn <- Fl*4 - (100+50*(Fl_attendants-1))
  if (MC > 0) MC_earn <- MC*4 - (100+50*(MC_attendants-1))
  return(list(Fl_earn, MC_earn))
}
earn_Fl <- NULL
earn_MC <- NULL
for (i in 7:21) {
  earn <- compute_earnings2(sel_date, i, 3, 2)
  earn_Fl <- c(earn_Fl, earn[[1]])
  earn_MC <- c(earn_MC, earn[[2]])
}
earn_Fl
earn_MC
sum(earn_Fl)
sum(earn_MC)
```

## Question 4

a. Now scale your calculations up for the full month of March (assuming that you are open 7 days a week). How much do you earn at most, and at worst? At both locations, assuming the same weather conditions as in the given data.

```{r}
earn_Fl <- NULL
earn_MC <- NULL
for (j in 1:31) {
  sel_date <- ymd("2013-03-01") + days(j)
  for (i in 7:20) {
    earn <- compute_earnings(sel_date, i, 3, 2)
    earn_Fl <- c(earn_Fl, earn[[1]])
    earn_MC <- c(earn_MC, earn[[2]])
  }
}
sum(earn_Fl)
sum(earn_MC)
```

b. During the month your coffee machine breaks, and you need to buy a new one. The new one costs $20000. Can you afford it?

`Yes, at Melbourne Central, with the month's earnings, but not at Flinders St which didn't make enough in the month.`

## TURN IN 

- Your `.Rmd` file
- Your Word (or pdf) file that results from knitting the Rmd.
- Make sure your group members are listed as authors, one person per group will turn in the report
- DUE: Wednesday after the lab, by 7am, loaded into moodle



