---
title: "ETC2420/5242 <br> Linear regression in R"
author: Earo Wang <br> <earo.wang@gmail.com>
date: Lab 05
---

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
opts_chunk$set(warning = FALSE, message = FALSE, 
			   fig.path = 'figure/', cache.path = 'cache/', 
			   fig.align = 'center', dev = 'png', fig.width = 10, fig.height = 6, 
			   fig.show = 'hold', cache = TRUE, external  =  TRUE,
         dev = 'Cairo_png')
read_chunk("lab05.R")
```

# Exploratory data analysis (EDA) with R

> The greatest value of a picture is when it forces us to notice what we never expected to see. -- J. W. Tukey (1977)

* Load the necessary packages in the beginning

```{r library}
```

* Take a glimpse at the dataset `gapminder`

```{r glimpse}
```

## Data visualisation

* It's difficult to process numerical information in raw form.
* That's why statistical graphics come to play.

```{r vis-country}
```

## Data modelling

<ul> 
<li>Transform</li>

```{r year1950}
```

<li>Start with Australia</li>

```{r au}
```

<li>Fit a simple regression for Australia

$$\widehat{\mathrm{lifeExp}}_{i} = \hat{\beta}_0 + \hat{\beta}_1 \times \mathrm{year1950}_{i}$$

<ol>
<li>Graphically</li>

```{r au-plot}
```

<li>Statistically</li>

```{r au-lm}
```

<li>Prediction</li>

```{r au-predict}
```

<li>Diagnostics

```{r tidy-lm}
```

<ul>
<li>Hat matrix $H$</li>
</li>
</li>
</ul>
</ol>
</ul>
$$H = X(X'X)^{-1}X' $$
$$\hat{y} = Hy = X\hat{\beta}$$

# Fitting many models at a time (using `purrr`)

There exists a more **elegant** way to fit many models than a `for loop`.

## Data structures --- lists

```{r list}
```

## Nesting the data into a list

```{r nest-data}
```

## Mapping a list of the data to `lm`

```{r map-lm}
```

## Unnesting a list of `lm` back to data frame

```{r map-tidy}
```

## Making a tidy report

```{r report}
```

## Visualising data

* Static

```{r ggplot-colour}
```

* Interactive

```{r plotly, eval = FALSE}
```

<iframe src="inter.html" height = "85%" width = "90%" align = "middle"></iframe>

## Better models?

```{r diag}
```

# Misc

## Online R resources

* [R for data science](http://r4ds.had.co.nz/lists.html) 
* [Managing many models](http://wombat2016.org/slides/hadley.pdf) given by Hadley Wickham
  at Wombat 2016, along with its [youtube video](https://www.youtube.com/watch?v=hRNUgwAFZtQ)

## Consultation hours

* Earo
    * **Time**: 13:30 to 15:00 on Thursday
    * **Venue**: Level 11/W1106, Menzies Building
* Nathaniel
    * **Time**: 14:00 to 15:00 on Tuesday
    * **Venue**: Level 11/W1106, Menzies Building

<meta name="copyright" content="LICENSE: CC BY-NC 3.0 US" />
