---
title: "Introduction to R"
author: "Nathaniel Tomasetti"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

# What is R?

R is the language of data analysis.

-It works with **objects** and **functions**.

-Its basic functions are similar to what you can do in Excel.

-But it has almost every statistical and data analytics tool available.

-It easily allows people to create their own functions and share them with the world.

-It is free, available on any operating system and growing in popularity in business.

-It can be hard to learn at first, but it gets easier the more you use it.

---

class:middle, inverse, center
# Objects in R

---

Objects allow things to be created and stored.

--

They can be variables,
```{r}
x <- 5
x + 1
```
--

or they can be vectors, matrices, and entire datasets.
```{r}
y <- c(1, 4, 2, 6, 5, 0)
y
z <- matrix(y, nrow = 2, ncol = 3)
z
```
--

An object can be almost everything!


---

class:middle, inverse, center
#Functions in R

---

Functions are a set of instructions to turn an input into an output.

```{r}
myMean <- function(x){
  N <- length(x)
  mean <- sum(x) / N
  return(mean)
}
x <- c(4, 1, 6, -3, 7, 1)
myMean(x)
```

--

It's easy to give them multiple inputs.

```{r}
addAndDouble <- function(x, y){
  2 * (x + y)
}
addAndDouble(5, 2)
```

---

R comes with many built in functions
```{r}
x <- c(4, 1, 6, -3, 7, 1)
mean(x)
```
--
```{r}
max(x)
```
--
```{r}
sd(x)
```
--
```{r}
cor(x, c(0, -5, 1, 3, 11, 2))
```
--
```{r}
sort(x)
```

---

R can do all of the regression you would do in eViews, Excel, Stata, SAS etc.

```{r}
x <- rnorm(1000)
y <- 2 + 3*x + rnorm(1000, sd = 0.1)
linReg <- lm(y ~ x)
summary(linReg)
```

---
class: middle, inverse, center

#Packages and Datasets

---

There are thousands of packages people have written with extra functions and data.

The dataset economics is part of the ggplot2 package and the function glimpse is part of the dplyr package. Install these once per computer.

```{r, eval=FALSE}
install.packages("ggplot2")
install.packages("dplyr")
```

--

Then use library(ggplot2) and library(dplyr) to load their contents into R.

```{r, eval=FALSE}
library(ggplot2)
library(dplyr)
data(economics, package="ggplot2")

```

--

Now we have access to everything in dplyr and ggplot2 until you close R!
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
data(economics, package="ggplot2")
```

Now we can look at this economics data

```{r}
glimpse(economics)
```

Alternatively, you can use any of these to get an idea about what the dataset looks like

```{r, eval=FALSE}
head(economics)
tail(economics)
summary(economics)
str(economics)
hist(economics$pce)
plot(economics$date, economics$pop)
```

---

If you're ever stuck on how to use a function, you can get help with ?

```{r, eval=FALSE}
?mean
?ggplot
```


There's also tonnes of online resources available,


- https://stackoverflow.com/ is a great resource for programming help

- https://cran.r-project.org/ has documentation on almost every function

- https://www.datacamp.com/ has nice interactive courses on learning R.

